---
title: "241102_ProteomicsModels"
author: "Albert Groenendijk"
date: "2024-11-02"
output: html_document
---

Load libraries
```{r}
library(tidyverse)
library(survival)
library(ggplot2)
library(riskRegression) # for time-dependent AUC calculation
library(randomForestSRC)
library(glmnet) # also loaded as part of riskRegression when you call GLMnet()
library(gbm)
```

Define directories
```{r}
inputdir <- "HIV_controllers_ATHENA/AthenaOlinkClinicalCombined_ClinicalPredictionModels/Input/"
scriptsdir <- "HIV_controllers_ATHENA/AthenaOlinkClinicalCombined_ClinicalPredictionModels/Scripts/"
outputdir <- "HIV_controllers_ATHENA/AthenaOlinkClinicalCombined_ClinicalPredictionModels/Output/"
```

Define generic objects
```{r}
`%notin%` <- Negate(`%in%`)
```

Set random seed
```{r}
set.seed(101)
```

Load datasets
```{r}
train <- list(
  VL4K = readRDS(paste0(outputdir,"proteomics_VL4K_train.rds")),
  HYBRID = readRDS(paste0(outputdir,"proteomics_HYBRID_train.rds")),
  CD4 = readRDS(paste0(outputdir,"proteomics_CD4_train.rds")))
test <- list(
  VL4K = readRDS(paste0(outputdir,"proteomics_VL4K_test.rds")),
  HYBRID = readRDS(paste0(outputdir,"proteomics_HYBRID_test.rds")),
  CD4 = readRDS(paste0(outputdir,"proteomics_CD4_test.rds")))

featurelist_olink <- readRDS(paste0(outputdir, "featurelist_olink.rds"))
```

# Define the clinical and proteomic predictors

Clinical predictors are prefilterd to be only the 4 most important ones:

- Log10Vl
- baseCD4sqrt
- MALESEX
- AGE

For the proteomic predictors, every dataframe only contains proteins with P <0.05.

Make a proteomic selection with the main proteins that are reported in Figure 3C-F in the paper.
```{r}
do_not_consider <- c("SMOKING_NEVER",
                     "baseHCVever","baseHBVever","baseHBV","baseHCV",
                     "baseBMI","time_since_2000",
                     "IDU", "ORIGIN_NONWESTERN")
clinicalpredictors <- c("baseVL_log10", "baseCD4_sqrt", "MALESEX", "AGE")
proteomic_selection <- c("ICAM3", "CD6", "LY9", "CRTAM",
                         "SH2D1A",  "C1QL2",  "CNGB3")
shared_proteins <- Reduce(rbind, featurelist_olink)  %>%
  group_by(PROTEIN) %>%
  filter(n() == 3) %>%
  ungroup() %>% pull(PROTEIN)
outcomes <- c("VL4K", "HYBRID", "CD4")

# Filter the columns and change factorvars to class factor
factor_vars <- c("MALESEX")
for (outcome in outcomes){
  
  train[[outcome]] <- train[[outcome]] %>% 
    select(-c(all_of(do_not_consider)))
  
  test[[outcome]] <- test[[outcome]] %>% 
    select(-c(all_of(do_not_consider)))
  
  for (var in factor_vars) {
    train[[outcome]][[var]] <- factor(train[[outcome]][[var]])
    test[[outcome]][[var]] <- factor(test[[outcome]][[var]],
      levels = levels(train[[outcome]][[var]]))
    # Aligns factor levels with training data
  }
}
```

## Optionally, run data integrity checks
```{r}
checkdata = T

if (checkdata){
  # print the dimensions
  for (i in c(1:3)){
    print("train dimensions (rows, columns):")
    print(dim(train[[i]]))
    print("test dimensions (rows, columns):")
    print(dim(test[[i]]))
  }
  
  # Check the number with events
  for (i in c(1:3)){
    print("train with event:")
    print(summary.factor(train[[i]]$status))
    print("test with event:")
    print(summary.factor(test[[i]]$status))
  }
  
  # Colnames between train and test
  for (i in c(1:3)){
    print(paste(outcomes[i], "column check:"))
    columncheck <- all(names(train[[i]] == names(test[[i]])))
    print(paste("Are colnames the same between train and test?",columncheck))
  }
  
  # Are all proteins with P<0.05 present?
  for (i in c(1:3)){
    inputproteins <- featurelist_olink[[outcomes[i]]] %>% filter(RAW_P < 0.05) %>% pull(PROTEIN) 
    columnsnotpresent <- inputproteins %notin% colnames(train[[i]])
    print(paste("No. of missing protein columns:", 
                sum(columnsnotpresent)))
  }
  
  # Are columns present that should not be there
  for (i in c(1:3)){
    columnspresent <- do_not_consider %in% colnames(train[[i]])
    print(paste("No. of columns that should not be considered:",
                sum(columnspresent)))
  }
  
  # Are all factors of class 'factor'
  for (i in c(1:3)){
    for (var in factor_vars){
      checkfactorcolumns <- class(train[[1]][[var]]) == "factor"
      print(paste(var, "is a factor:", checkfactorcolumns))
    }
  }
  
  # Identify missingness
  for (i in c(1:3)){
    na_counts <- sapply(train[[i]], function(x) sum(is.na(x)))
    if (sum(na_counts) > 1){
      print("There are missing in train")
      print(na_counts)
    }else{ print("No NAs in train") }

    na_counts <- sapply(test[[i]], function(x) sum(is.na(x)))
    if (sum(na_counts) > 1){
      print("There are missing in test")
      print(na_counts)
    }else{ print("No NAs in test") }
  }
}
```

## Random train-test split to verify
```{r}
outcomes <- c("VL4K","HYBRID", "CD4")
set.seed(123)
randomtrain <- list()
randomtest <- list()
for (outcome in outcomes){
  fulldata <- rbind(train[[outcome]], test[[outcome]])
  randomtrain[[outcome]] <- fulldata %>% slice_sample(prop = 0.7)
  randomtest[[outcome]] <- anti_join(fulldata, randomtrain[[outcome]], by = "ID")
}


```


# Train the models
## simple Cox regression

Cox regression is very sensitive to overfitting. We take here a very small number of variables.

```{r}
# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")

outcome <- outcomes[1]


# Define features
## All features
features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))
formula_model_medium <- as.formula(paste("Surv(time, status) ~", paste(c(clinicalpredictors, shared_proteins), collapse = " + ")))
formula_model_small <- as.formula(paste("Surv(time, status) ~", paste(c(clinicalpredictors, proteomic_selection), collapse = " + ")))

# Train the cox model
## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
if ("CD6" %notin% colnames(train[[outcome]])){
  dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
}else{
  dataset_train <- train[[outcome]]
  dataset_test <- test[[outcome]]
}

coxmodel <- coxph(formula_model_small, dataset_train, x = TRUE)

times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(time, status) ~ 1"))
score_result <- Score(list("Cox" = coxmodel),
        formula = survformula,
        data = dataset_test,
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
AUC_overtime$outcome <- "outcome"
print(AUC_overtime)

#--------------------------------------------
# Redo the model with random train-test split
#--------------------------------------------
## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
if ("CD6" %notin% colnames(train[[outcome]])){
  dataset_randomtrain <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  dataset_randomtest <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
}else{
  dataset_randomtrain <- train[[outcome]]
  dataset_randomtest <- test[[outcome]]
}

coxmodel <- coxph(formula_model_small, dataset_randomtrain, x = TRUE)

times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(time, status) ~ 1"))
score_result <- Score(list("Cox" = coxmodel),
        formula = survformula,
        data = dataset_randomtest,
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
AUC_overtime$outcome <- "outcome"
print(AUC_overtime)
```

## Elastic net cox regression

This is less sensitive to noise & overfitting.
alpha = 1 for lasso, alpha = 0 for ridge

Ridge (L2), alpha = 0: Useful when you suspect all variables may be relevant but need regularization.
Lasso (L1), alpha = 1: Helps with feature selection by pushing less informative coefficients to zero, simplifying the model.


Test the code
```{r}
outcome <- outcomes[1]

# Define the formula
formula_model_small <- as.formula(paste("Surv(time, status) ~", paste(c(clinicalpredictors, proteomic_selection), collapse = " + ")))

## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
if ("CD6" %notin% colnames(train[[outcome]])){
  dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
}else{
  dataset_train <- train[[outcome]]
  dataset_test <- test[[outcome]]
}

penalizedmodel_lasso <- GLMnet(formula_model_small, data = dataset_train, alpha = 1, lambda = NULL, cv = TRUE, nfolds = 10, standardize = TRUE)
```

Find optimal by crossvalidation
```{r}
# Define a sequence of alpha values to test
alpha_values <- seq(0, 1, by = 0.1)  # Adjust step size as needed

# Initialize a data frame to store results
elasticnet_cvresults <- data.frame(alpha = numeric(), lambda_min = numeric(), cvm_min = numeric(), outcome = character())

for (outcome in outcomes){
  
  ## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
  if ("CD6" %notin% colnames(train[[outcome]])){
    dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
    dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  }else{
    dataset_train <- train[[outcome]]
    dataset_test <- test[[outcome]]
  }
  
  # Loop through each alpha value
  for (alpha in alpha_values) {
    # Perform cross-validation for each alpha
    set.seed(123)  # For reproducibility
    
    penalizedmodel_balanced <- GLMnet(formula_model_small, data = dataset_train, alpha = alpha, lambda = NULL, cv = TRUE, nfolds = 10, standardize = TRUE)
  
    # Store the results: best lambda and corresponding CV error
    elasticnet_cvresults <- rbind(
      elasticnet_cvresults, 
      data.frame(alpha = alpha,
                 lambda_min = penalizedmodel_balanced$fit$lambda.min,
                 cvm_min = min(penalizedmodel_balanced$fit$cvm),
                 outcome = outcome))
  }
}
elasticnet_cvresults

# Find the optimal alpha based on minimum CV error
optimal_alpha_value <- elasticnet_cvresults %>% group_by(outcome) %>% filter(cvm_min == min(cvm_min)) %>% select(alpha, outcome) %>%  pull(alpha, outcome)

print(paste0('optimal alpha for glmnet: ',optimal_alpha_value))
```

Try both prefiltered features and full featurelist
```{r}
outcome <- outcomes[1]

  ## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
  if ("CD6" %notin% colnames(train[[outcome]])){
    dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
    dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  }else{
    dataset_train <- train[[outcome]]
    dataset_test <- test[[outcome]]
  }

penalizedmodel_lasso <- GLMnet(formula_model_small, data = dataset_train, alpha = 0, lambda = NULL, cv = TRUE, nfolds = 10, standardize = TRUE)

times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(time, status) ~ 1"))
score_result <- Score(list("Penalized" = penalizedmodel_lasso),
        formula = survformula,
        data = dataset_test,
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)

features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))

penalizedmodel_lasso <- GLMnet(formula_model, data = dataset_train, alpha = 1, lambda = NULL, cv = TRUE, nfolds = 10, standardize = TRUE)

times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(time, status) ~ 1"))
score_result <- Score(list("Penalized" = penalizedmodel_lasso),
        formula = survformula,
        data = dataset_test,
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```

## Gradient boosted machines

The disadvantage of using gbm() with Score() is that the formula in gbm() must be typed statically, which is very long in this case.
```{r}
outcome <- outcomes[1]

features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))

formula_model
formula_model_small

## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
if ("CD6" %notin% colnames(train[[outcome]])){
  dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
}else{
  dataset_train <- train[[outcome]]
  dataset_test <- test[[outcome]]
}

# Fit a GBM model with survival data
gbm_model <- gbm(
    formula = Surv(time, status) ~ 
    MALESEX + baseVL_log10 + AGE + baseCD4_sqrt + 
    GZMH + ITGB1BP2 + ADH4 + EPHX2 + ADAMTS16 + CA5A + ACOX1 + 
    MCFD2 + FCRL1 + ADGRG2 + ROR1 + CA13 + CCDC80 + SFTPD + PRKAR1A + 
    CDHR5 + IL6ST + GDF2 + NTRK2 + LGALS1 + CDH6 + ADAMTS13 + 
    ICAM3 + SELE + CCN3 + LILRB2 + ITGB2 + AMY2A + SEMA7A + XG + 
    ADGRE5 + PDGFRB + NCAM1 + OSMR + RARRES2 + NRP1 + ADA2 + 
    IGFBP3 + DPP4 + HSPG2 + ENAH + IL17A + IL5 + IL12RB1 + MICB_MICA + 
    BID + SCRN1 + PTH1R + ANXA11 + SLAMF7 + IL32 + RAB6A + ITGA6 + 
    FCRL6 + SULT2A1 + SH2D1A + KLRB1 + PRDX5 + KLRD1 + CD79B + 
    FCRL2 + ADAM23 + CD40LG + IDS + GZMA + CD6 + CD160 + LY9 + 
    MZB1 + ENPP5 + CD58 + HSPA1A + CD48 + CCL4 + CXCL10 + CCL17 + 
    CXCL9 + TNFRSF13B + CXCL3 + CCL24 + LGALS9 + CD276 + MEGF10 + 
    ADGRE2 + CCL22 + PRELP + CA11 + ADCYAP1R1 + NTF4 + NPTN + 
    GFER + NAMPT + SUGT1 + FCGR2B + GSTA3 + LTBP3 + FCRLB + RAD23B + 
    PSRC1 + HGS + ABL1 + TBC1D23 + MPI + GFRA2 + CALB1 + CDKN1A + 
    KAZALD1 + IGF1R + CDNF + KLK13 + USO1 + RBP5 + VAT1 + CDKN2D + 
    CD207 + DDAH1 + LAG3 + CBLN4 + PDCD1 + RTN4R + FXN + ITGB7 + 
    CD5 + PRDX6 + METAP2 + ITGAV + CLMP + DAB2 + CD27 + PPY + 
    C4BPB + ATOX1 + FGFBP1 + ITGB5 + KLB + TST + C19orf12 + CRADD + 
    PPP3R1 + ABHD14B + TNFRSF9 + FLRT2 + ASGR1 + GGT5 + ADAM22 + 
    CRTAM + CHMP1A + FKBP4 + ROBO2 + SCARF2 + LAMP2 + CD8A + 
    CXCL13 + TNFRSF8 + THY1 + CTSS + NTRK3 + SCARA5 + CGA + SPINK5 + 
    WARS + IL17RA + NCAM2 + CNTN4 + CD99L2 + CD99 + FCRL5 + NOP56 + 
    EXOSC10 + CSF2 + CD2 + SHD + IGHMBP2 + C1GALT1C1 + SIGLEC8 + 
    CCER2 + HSBP1 + KHK + NIT1 + GATD3 + ASPN + NIT2 + NGFR + 
    BPIFB2 + PGD + RAB27B + LCP1 + COL15A1 + PTPRC + POSTN + 
    LATS1 + TNFSF9 + KLRF1 + PTP4A3 + C1QL2 + AKAP12 + SCGB3A1 + 
    ADH1B + C1QTNF9 + RIDA + CD72 + GHR + TNFRSF17 + NEXN + C9 + 
    F10 + AHSG + AMY1A_AMY1B_AMY1C + B2M + SERPINA5 + IGLC2 + 
    C1R + C1S + ORM1 + C3 + DNLZ + CNGB3 + TMCO5A + ERC2 + CABP2 + 
    CAMLG + MDM1 + CLGN + EDF1 + RGCC + IDO1 + DSCAM + CDK5RAP3 + 
    STXBP1 + TSC22D1 + RGS10 + SUOX + LCN15 + DXO + PLXDC2 + 
    PNLIPRP1 + PLCB2 + KLRK1 + KIAA0319 + PALM2 + GSTT2B + DLG4 + 
    DHRS4L2 + MINK1 + TXNDC9 + PLSCR3 + VWA5A + LYZL2 + TIGIT + 
    REG3G + ZNRF4 + AMOTL2 + ACRBP + FAM171B + TADA3 + TARS1 + 
    CHCHD6 + AP3B1 + IFNAR1 + GPHA2 + NFU1,  # Survival formula
    data = dataset_train,
    distribution = "coxph",    # Cox model for survival
    n.trees = 1000,             # Number of trees
    interaction.depth = 3,     # Tree depth
    shrinkage = 0.01,          # Learning rate
    n.minobsinnode = 15,       # Minimum observations in terminal nodes
    bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
)

# Use Score function with the GBM survival model
times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(time, status) ~ 1"))
score_result <- Score(
    list("GBM Model" = gbm_model), 
    formula = survformula,   # Survival formula for Score function
    data = dataset_test,                        # Original data with survival info
    times = times,                # Time points to evaluate the survival model
    metrics = "AUC",                  # Metric for model evaluation
    summary = "risks"                     # Summary option for integrated Brier score
)

# Print results
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```

## RSF model
Non-parametric machine learning approach.

The rfsrc() function offers many parameters for fine-tuning:
* mtry: Number of variables randomly selected at each split
* nodesize: Minimum size of terminal nodes
* splitrule: Rule used for splitting nodes

```{r}
outcomes <- c("VL4K","HYBRID", "CD4")

outcome <- outcomes[1]

features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))

## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
if ("CD6" %notin% colnames(train[[outcome]])){
  dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
    test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
}else{
  dataset_train <- train[[outcome]]
  dataset_test <- test[[outcome]]
}

rsfmodel <- rfsrc(formula_model, data = dataset_train, ntree = 3000, cv.fold = 10, nodesize = 15, mtry = 4, splitrule = "logrankscore", nsplit = 10)


times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(time, status) ~ 1"))
score_result <- Score(list("RSF" = rsfmodel),
        formula = survformula,
        data = dataset_test,
        metrics = c("auc"),
        summary = "risks",
        times = times,
        plots = "ROC")

AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
ROC_overtime <- score_result$ROC$plotframe
ROC_overtime$type <- "test"
AUCROC_overtime <- list(AUC_overtime = AUC_overtime, ROC_overtime = ROC_overtime)
print(AUCROC_overtime)
AUC_overtime <- AUCROC_overtime$ROC_overtime

score_result$ROC$plotframe


roc_data_5yr <- subset(score_result$ROC$plotframe, times == 5*365)




# Plot using ggplot2
ggplot(roc_data_5yr, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  labs(
    title = "ROC Curve at 5-Year Risk",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()



AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```

## Display the relevance of each predictor

```{r}
meandecreaseaccuracy = list()

obtain_meandecreaseaccuracy <- function(outcome){
  ## Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
  if ("CD6" %notin% colnames(train[[outcome]])){
    dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
    dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  }else{
    dataset_train <- train[[outcome]]
    dataset_test <- test[[outcome]]
  }

  # train the RSF, random survival forest
  features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
  formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))
  rsfmodel <- rfsrc(formula_model, data = dataset_train, ntree = 3000, cv.fold = 10, nodesize = 15, mtry = 4, splitrule = "logrankscore", nsplit = 10)

importance_df <- tibble(
  Outcome = outcome,
  Variable = names(vimp(rsfmodel)$importance),
  MeanDecreaseAccuracy = vimp(rsfmodel)$importance
  ) 

meandecreaseaccuracy[[outcome]] <- importance_df
return(meandecreaseaccuracy)
}

obtain_meandecreaseaccuracy(outcomes[3])


# Make a list of all results
for (outcome in outcomes){
  meandecreaseaccuracy <- obtain_meandecreaseaccuracy(outcome)
}

# Perform cleaning operations on the list
for (outcome in outcomes){
  meandecreaseaccuracy[[outcome]] <- meandecreaseaccuracy[[outcome]] %>%
    mutate(
      Outcome = factor(
        Outcome, 
        levels = c("VL4K", "HYBRID", "CD4"),
        labels = c("Virological loss", 
                   ">1000 copies/mL increase", 
                   "CD4 <500 cells/mm3")))
}

# Create the ggplot
var_importance_plot <- function(outcome_index){
  i <- outcome_index
  data <- meandecreaseaccuracy[[outcomes[i]]] %>% 
  arrange(desc(MeanDecreaseAccuracy)) %>% head(20)
  
  fillcolor <- c("#5ec962","orange","darkred")[i]
  g <- ggplot(data, aes(x = reorder(Variable, MeanDecreaseAccuracy, .desc = TRUE), y = MeanDecreaseAccuracy)) +
  geom_bar(stat = "identity", fill = fillcolor, width = 0.5) +
  coord_flip() +
  labs(title = "Variable Importance in Random Survival Forest",
       x = "Variables",
       y = "Mean Decrease in Accuracy") +
  theme_bw() +
  theme(strip.placement = "outside",
     strip.text = element_text(angle = 0, size = 8),
     strip.background = element_blank(),
     axis.text = element_text(size = 8, color = "black"),
     axis.title.y = element_blank(),
     axis.title.x = element_text(size = 10),
     plot.title = element_text(size = 12, face = "bold")) + 
    
    annotate("text", x = 5, y = max(data$MeanDecreaseAccuracy), 
             angle = 90, size = 8 / .pt, vjust = -1,
             label = meandecreaseaccuracy[[outcomes[i]]]$Outcome[1]) 
  
  print(g)
}

for (i in c(1:length(outcomes))){
  ggsave(paste0(outputdir, "VariableImportanceRSF_Proteomics_", outcomes[i], ".pdf"),var_importance_plot(i),width = 21/3, height = 29.7/2.75, units = "cm")

}
```

Define a function for the gbm where the formula is statically typed
```{r}
# outcome <- outcomes[1]
# features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
# formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))
# formula_model

obtain_gbmmodel <- function(outcome, dataset_train){
  # Fit a GBM model with survival data
  ## Formula for gbm() has to be statically typed
  if (outcome == "VL4K"){
    gbm_model <- gbm(
        formula = Surv(time, status) ~ 
          MALESEX + baseVL_log10 + AGE + baseCD4_sqrt + 
          GZMH + ITGB1BP2 + ADH4 + EPHX2 + ADAMTS16 + CA5A + ACOX1 + 
          MCFD2 + FCRL1 + ADGRG2 + ROR1 + CA13 + CCDC80 + SFTPD + PRKAR1A + 
          CDHR5 + IL6ST + GDF2 + NTRK2 + LGALS1 + CDH6 + ADAMTS13 + 
          ICAM3 + SELE + CCN3 + LILRB2 + ITGB2 + AMY2A + SEMA7A + XG + 
          ADGRE5 + PDGFRB + NCAM1 + OSMR + RARRES2 + NRP1 + ADA2 + 
          IGFBP3 + DPP4 + HSPG2 + ENAH + IL17A + IL5 + IL12RB1 + MICB_MICA + 
          BID + SCRN1 + PTH1R + ANXA11 + SLAMF7 + IL32 + RAB6A + ITGA6 + 
          FCRL6 + SULT2A1 + SH2D1A + KLRB1 + PRDX5 + KLRD1 + CD79B + 
          FCRL2 + ADAM23 + CD40LG + IDS + GZMA + CD6 + CD160 + LY9 + 
          MZB1 + ENPP5 + CD58 + HSPA1A + CD48 + CCL4 + CXCL10 + CCL17 + 
          CXCL9 + TNFRSF13B + CXCL3 + CCL24 + LGALS9 + CD276 + MEGF10 + 
          ADGRE2 + CCL22 + PRELP + CA11 + ADCYAP1R1 + NTF4 + NPTN + 
          GFER + NAMPT + SUGT1 + FCGR2B + GSTA3 + LTBP3 + FCRLB + RAD23B + 
          PSRC1 + HGS + ABL1 + TBC1D23 + MPI + GFRA2 + CALB1 + CDKN1A + 
          KAZALD1 + IGF1R + CDNF + KLK13 + USO1 + RBP5 + VAT1 + CDKN2D + 
          CD207 + DDAH1 + LAG3 + CBLN4 + PDCD1 + RTN4R + FXN + ITGB7 + 
          CD5 + PRDX6 + METAP2 + ITGAV + CLMP + DAB2 + CD27 + PPY + 
          C4BPB + ATOX1 + FGFBP1 + ITGB5 + KLB + TST + C19orf12 + CRADD + 
          PPP3R1 + ABHD14B + TNFRSF9 + FLRT2 + ASGR1 + GGT5 + ADAM22 + 
          CRTAM + CHMP1A + FKBP4 + ROBO2 + SCARF2 + LAMP2 + CD8A + 
          CXCL13 + TNFRSF8 + THY1 + CTSS + NTRK3 + SCARA5 + CGA + SPINK5 + 
          WARS + IL17RA + NCAM2 + CNTN4 + CD99L2 + CD99 + FCRL5 + NOP56 + 
          EXOSC10 + CSF2 + CD2 + SHD + IGHMBP2 + C1GALT1C1 + SIGLEC8 + 
          CCER2 + HSBP1 + KHK + NIT1 + GATD3 + ASPN + NIT2 + NGFR + 
          BPIFB2 + PGD + RAB27B + LCP1 + COL15A1 + PTPRC + POSTN + 
          LATS1 + TNFSF9 + KLRF1 + PTP4A3 + C1QL2 + AKAP12 + SCGB3A1 + 
          ADH1B + C1QTNF9 + RIDA + CD72 + GHR + TNFRSF17 + NEXN + C9 + 
          F10 + AHSG + AMY1A_AMY1B_AMY1C + B2M + SERPINA5 + IGLC2 + 
          C1R + C1S + ORM1 + C3 + DNLZ + CNGB3 + TMCO5A + ERC2 + CABP2 + 
          CAMLG + MDM1 + CLGN + EDF1 + RGCC + IDO1 + DSCAM + CDK5RAP3 + 
          STXBP1 + TSC22D1 + RGS10 + SUOX + LCN15 + DXO + PLXDC2 + 
          PNLIPRP1 + PLCB2 + KLRK1 + KIAA0319 + PALM2 + GSTT2B + DLG4 + 
          DHRS4L2 + MINK1 + TXNDC9 + PLSCR3 + VWA5A + LYZL2 + TIGIT + 
          REG3G + ZNRF4 + AMOTL2 + ACRBP + FAM171B + TADA3 + TARS1 + 
          CHCHD6 + AP3B1 + IFNAR1 + GPHA2 + NFU1,  # Survival formula
        data = dataset_train,
        distribution = "coxph",    # Cox model for survival
        n.trees = 1000,             # Number of trees
        interaction.depth = 3,     # Tree depth
        shrinkage = 0.01,          # Learning rate
        n.minobsinnode = 15,       # Minimum observations in terminal nodes
        bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
    )
  }else if (outcome == "HYBRID"){
    gbm_model <- gbm(
        formula = Surv(time, status) ~ 
          MALESEX + baseVL_log10 + AGE + baseCD4_sqrt + 
          TSLP + GZMH + CTSH + EDIL3 + ZBTB17 + ADAMTS16 + ACOX1 + 
          PRSS27 + SFTPD + CDHR5 + KITLG + IL6ST + GDF2 + LGALS1 + 
          IL19 + TNFSF13B + ICAM3 + ITGB2 + PLXNB2 + UMOD + XG + ADGRE5 + 
          PDGFRB + PRSS2 + IGFBP3 + HSPG2 + NBN + CLEC4A + CD4 + IL5RA + 
          GZMB + LSP1 + PSPN + SH2D1A + BTN3A2 + KLRB1 + CD244 + KLRD1 + 
          IL16 + LAMP3 + IL18R1 + CXCL17 + GZMA + CD6 + CD160 + LY9 + 
          SCGB1A1 + SIGLEC1 + SMOC2 + CD48 + CCL4 + IL1RN + CXCL9 + 
          ANGPTL4 + LGALS9 + MEGF10 + ADGRE2 + PRELP + ADCYAP1R1 + 
          UXS1 + DTX3 + GFRA2 + ALPP + LAG3 + PDCD1 + ITGB7 + CD5 + 
          SMOC1 + PRTG + FGF23 + CD27 + PPY + KLB + IL34 + NDRG1 + 
          NOS3 + S100A16 + DPEP1 + PPP3R1 + SKAP1 + LY96 + PLA2G10 + 
          CRTAM + CSF2RA + TNR + EZR + CD8A + CDH15 + SEMA4D + CXCL13 + 
          CXCL11 + TNFRSF8 + NTRK3 + IL17RA + NCAM2 + GNLY + CD109 + 
          CNTN4 + NOP56 + EXOSC10 + TPR + ENOX2 + DDA1 + CD2 + LRCH4 + 
          SNU13 + SART1 + EDEM2 + INPP5D + ANXA2 + ROBO4 + LIPF + ADAMTSL4 + 
          GIP + RECK + AAMDC + ASPN + TCOF1 + TIMP2 + CPXM2 + BPIFB2 + 
          TPK1 + LCP1 + COL3A1 + PTPRC + POSTN + ITGAL + ZBP1 + TNFSF9 + 
          KLRF1 + NECTIN1 + C1QL2 + SCGB3A1 + DIPK2B + MXRA8 + GIMAP7 + 
          KLK7 + RLN2 + RNASE4 + APOL1 + CFHR2 + B2M + IGLC2 + C1R + 
          MBL2 + KLF4 + CNGB3 + IMPG1 + PMS1 + PPT1 + TPSD1 + SAFB2 + 
          REEP4 + DNPEP + PNLIPRP1 + ZHX2 + KLRK1 + C7orf50 + YAP1 + 
          GBP1 + CEACAM20 + NEK7 + TIGIT + WDR46 + BTN1A1 + GLYR1 + 
          DYNLT3 + NENF + IL2RG + GPD1 + DYNC1H1 + NACC1 + RALY + CEACAM19 + 
          ATRAID + SWAP70 + BCL2 + LMNB1 + SMNDC1 + IFI30,  # Survival formula
        data = dataset_train,
        distribution = "coxph",    # Cox model for survival
        n.trees = 1000,             # Number of trees
        interaction.depth = 3,     # Tree depth
        shrinkage = 0.01,          # Learning rate
        n.minobsinnode = 15,       # Minimum observations in terminal nodes
        bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
    )
    
  }else if (outcome == "CD4"){
    gbm_model <- gbm(
        formula = Surv(time, status) ~ 
          MALESEX + baseVL_log10 + AGE + baseCD4_sqrt + 
          VSTM2L + MSTN + ACTA2 + DDC + MARCO + SORT1 + FBP1 + ICAM3 + 
          MFAP5 + FABP4 + FUCA1 + VASN + C2 + IGFBP6 + HSPG2 + JCHAIN + 
          IL1RL2 + AMBN + LRRN1 + FCRL3 + AMN + CNTNAP2 + CLEC4A + 
          PTX3 + LY75 + CLEC4G + KLRB1 + TNFRSF11A + LY9 + SCGB1A1 + 
          MZB1 + CCL4 + FST + RRM2 + CFC1 + RARRES1 + SERPINA9 + CALB1 + 
          GPA33 + OPTC + CCN1 + FLT1 + LAG3 + RTN4R + AMBP + CD300LF + 
          HSPB6 + CRISP2 + KLK6 + NELL1 + C4BPB + LYPD3 + KLK1 + CTSF + 
          KLB + NXPH1 + VSTM1 + NOS3 + PAK4 + WFIKKN1 + CRTAM + VWC2 + 
          FUT8 + NAAA + DKK1 + SEMA4D + THY1 + SMPD1 + SPINK5 + PIK3IP1 + 
          DSG2 + EFNA1 + FCRL5 + ARSA + MEGF11 + CPTP + GUK1 + ACTN2 + 
          SIGLEC8 + HRC + ROBO4 + CCER2 + LAMB1 + ECHDC3 + MAN2B2 + 
          BPIFB2 + FABP3 + GOT1 + CYTL1 + C1QL2 + AKAP12 + CST1 + C9 + 
          AMY1A_AMY1B_AMY1C + SHBG + CFD + C1R + C1S + SELENOP + ORM1 + 
          DTX2 + IDO1 + MCEE + DNPEP + CDH23 + VWA5A + TIGIT + CASP4 + 
          RNF43,  # Survival formula
        data = dataset_train,
        distribution = "coxph",    # Cox model for survival
        n.trees = 1000,             # Number of trees
        interaction.depth = 3,     # Tree depth
        shrinkage = 0.01,          # Learning rate
        n.minobsinnode = 15,       # Minimum observations in terminal nodes
        bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
    )
    
  }else{warning("Unrecognized outcome type")}
  
  return(gbm_model)
}

outcome <- outcomes[1]
gbm_model <- obtain_gbmmodel(outcome, dataset_train)
```

```{r}
# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")

# Specify time points of interest to measure AUC
times <- seq(2, 8) * 365

# Define function to get score result
## This function will be used in the function obtain_auc below
obtain_scores <- function(models_list, dataset, survformula, traintest){
  score_result <- Score(models_list,
        formula = survformula,
        data = dataset,
        metrics = "auc",
        summary = "risks",
        times = times,
        plots = "ROC")
  
  AUC_overtime <- score_result$AUC$score
  AUC_overtime$type <- traintest

  ROC_overtime <- score_result$ROC$plotframe
  ROC_overtime$type <- traintest
  
  AUCROC_overtime <- list(AUC_overtime = AUC_overtime, ROC_overtime = ROC_overtime)

  return(AUCROC_overtime)
}

# Define function to get the AUC over time
obtain_aucroc <- function(outcome, metric){
  if (metric %notin% c("AUC", "ROC")) {
    stop("Invalid metric! Metric must be either 'AUC' or 'ROC'")
  }
  if (outcome %notin% c("VL4K", "HYBRID", "CD4")) {
    stop("Invalid outcome! Metric must be one of: VL4K, HYBRID, CD4")
  }
  
  # Add CD6 and others for CD4 outcome, since these were not significant for CD4 outcome
  if ("CD6" %notin% colnames(train[[outcome]])){
    dataset_train <- train[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      train[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
    dataset_test <- test[[outcome]] %>% filter(!is.na(status)) %>% left_join(
      test[["VL4K"]] %>% select(ID, CD6, SH2D1A, CNGB3), by = "ID")
  }else{
    dataset_train <- train[[outcome]]
    dataset_test <- test[[outcome]]
  }

  # Train the cox model
  formula_model_small <- as.formula(paste("Surv(time, status) ~", paste(c(clinicalpredictors, proteomic_selection), collapse = " + ")))

  coxmodel <- coxph(formula_model_small, dataset_train, x = TRUE)
  
  # Define the formula for other models
  features <- train[[outcome]] %>% select(-c(time, status, ID)) %>% colnames()
  formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))

  # Train the elastic-net (balanced) penalized cox model
  elasticnetcox_model <- GLMnet(
    formula = formula_model,
    data = dataset_train,
    alpha = 1,
    lambda = NULL,
    cv = TRUE,
    nfolds = 10,
    standardize = TRUE)

  # train the RSF, random survival forest
  rsfmodel <- rfsrc(formula_model, data = dataset_train, ntree = 3000, cv.fold = 10, nodesize = 15, mtry = 4, splitrule = "logrankscore", nsplit = 10)

  # Fit a GBM model with survival data
  gbm_model <- obtain_gbmmodel(outcome = outcome, 
                               dataset_train = dataset_train)

  # Calculate AUC data over time for train and test
  survformula <- as.formula(paste("Surv(time, status) ~ 1"))
  model_list <- list("Cox" = coxmodel, 
                     "RSF" = rsfmodel,
                     "GBM" = gbm_model,
                     "ElasticNetCox" = elasticnetcox_model)
  auc_train <- obtain_scores(model_list, dataset_train, survformula, "train")$AUC_overtime
  auc_test <-  obtain_scores(model_list, dataset_test, survformula, "test")$AUC_overtime

  auc_combined <- rbind(auc_train, auc_test)
  auc_combined$outcome <- outcome
  
  roc_train <- obtain_scores(model_list, dataset_train, survformula, "train")$ROC_overtime
  roc_test <-  obtain_scores(model_list, dataset_test, survformula, "test")$ROC_overtime

  roc_combined <- rbind(roc_train, roc_test)
  roc_combined$outcome <- outcome

  if (metric == "AUC"){
    return(auc_combined)
  }else if (metric == "ROC"){
    return(roc_combined)
  }else{return(NA)}
}

# Apply the function to each outcome with a lambda function inside map_dfr
auc_data <- map_dfr(outcomes, ~obtain_aucroc(outcome = .x, metric = "AUC"))

roc_data <- map_dfr(outcomes, ~obtain_aucroc(outcome = .x, metric = "ROC"))
```

```{r}
# add nice labels for model and outcome
auc_data <- auc_data %>% mutate(
  model = factor(
    model, levels = c("Cox",
                      "RSF",
                      "ElasticNetCox",
                      "GBM"), 
    labels = c("Cox proportional hazards regression",
               "Random survival forest",
               "Elastic net cox regression",
               "Gradient boosted machines")),
  outcome = factor(outcome, 
    levels = c("VL4K", "HYBRID", "CD4"),
    labels = c("Virological loss", ">1000 copies/mL increase", "CD4 <500 cells/mm3")))

# Plot AUC over time with confidence intervals
## The plot should facetgrid on model on the x-axis and outcome on the y-axis. 
## outcome determines color and fill. type determines shape and linetype and group.
auc_plot <- ggplot(auc_data, aes(x = times, y = AUC, group = type)) +
  geom_line(aes(linetype = type, color = outcome), linewidth = 1) +
  geom_point(aes(shape = type, color = outcome), size = 2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = outcome), alpha = 0.2) +
  labs(x = "Time (years)", 
       y = "Area Under the Curve (AUC)", 
       title = "Prediction model performance over time (clinical & proteomic predictors)") +
  theme_bw() +
  scale_colour_manual(values=(c("#5ec962","orange","darkred"))) + 
  scale_fill_manual(values=(c("#5ec962","orange","darkred"))) + 
  scale_y_continuous(limits = c(0.30, 1.1),
    breaks = c(0.5, 0.75, 1.0)) +
  scale_x_continuous(limits = c(0, 9*365),
    breaks = seq(0, max(auc_data$times), by = 365),
    labels = function(x) round(x / 365, 1)
  ) +
  facet_grid(outcome ~ model) +
  theme(strip.placement = "outside",
     strip.text= element_text(angle = 0, size = 8),
     strip.background = element_blank(),
     axis.text = element_text(size = 10, color="black"),
     axis.title = element_text(size = 10, color="black", face = "bold"),
     title = element_text(size=12, color = "black"),
     legend.text = element_text(size = 10, color="black"),
     legend.title = element_blank(),
     legend.position = "bottom",
     plot.subtitle = element_text(size = 10, color= "black", face="italic"),
     #legend.key.height = unit(1.5,'cm'),
     panel.grid.minor.x = element_blank(),
     panel.spacing = unit(0, "mm")) +
  
  # keep the type legend, but remove the outcome legend
    guides(fill = "none", color = "none", shape = guide_legend(override.aes = list(size=3))) # Remove outcome legend

print(auc_plot)

ggsave(paste0(outputdir,"AUC_clinical_and_proteomics.pdf"), plot = auc_plot, width = 210, height = 297/2, units = "mm")
```
# ROC curves 

I want one plot with ROC curves, three outcomes
5-year timepoint, elasticnet data, test and train?

```{r}
summary.factor(roc_data$model)

roc_data_5yr <- roc_data %>% filter(times == 3*365) %>% 
  filter(model == "ElasticNetCox") %>% mutate(
    outcome = factor(outcome, 
      levels = c("VL4K", "HYBRID", "CD4"),
      labels = c("Virological loss", ">1000 copies/mL increase", "CD4 <500 cells/mm3")))

# Plot using ggplot2
roc_plot <- ggplot(roc_data_5yr, aes(x = FPR, y = TPR)) +
  facet_wrap(~ outcome, dir = "h") +
  geom_line(aes(color = outcome, linetype = type), linewidth = 0.65) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  scale_colour_manual(values=(c("#5ec962","orange","darkred"))) + 
  labs(
    title = "Receiver Operater Curve (ROC) at 3−Year Risk",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_bw()+
    theme(strip.placement = "outside",
     strip.text= element_text(angle = 0, size = 8),
     strip.background = element_blank(),
     axis.text = element_text(size = 10, color="black"),
     axis.title = element_text(size = 10, color="black", face = "bold"),
     title = element_text(size=12, color = "black"),
     legend.text = element_text(size = 10, color="black"),
     legend.title = element_blank(),
     legend.position = "bottom",
     plot.subtitle = element_text(size = 10, color= "black", face="italic"),
     #legend.key.height = unit(1.5,'cm'),
     panel.grid.minor.x = element_blank(),
     panel.spacing = unit(5, "mm")) +
  
  # keep the type legend, but remove the outcome legend
    guides(fill = "none", color = "none", shape = guide_legend(override.aes = list(size=3))) # Remove outcome legend

print(roc_plot)

ggsave(paste0(outputdir,"ROC_clinical_and_proteomics.pdf"), plot = roc_plot, width = 210, height = 297/3, units = "mm")
````





















