---
title: "251014_AgeMatchedPaired"
output: html_document
author: "Albert Groenendijk"
date: "2025-10-14"
---

A spin-off from the main notebook `AthenaOlink_LinearMixedModels_PairedSamples.Rmd`

The reviewer points out the absolute age differences between EC, persistent VC and transient VC.
To address this, we age-match the groups to EC  and redo the paired before/after comparisons with nlme. 

## MATCHING
### Summary of group sizes

* n=57 persistent VC who start ART
* n=36 Transient VC who start ART
* n=9 EC who start ART
* Additionally 19 EC and 23 VC are included who have paired samples

```{r echo=T, results='hide', warning=F,message=F}
working_dir <- c("HIV_controllers_ATHENA/AthenaOlink_LinearMixedModels_PairedSamples/")

library(FactoMineR)
library(dplyr)
library(magrittr)
library(lubridate)
library(table1)
library(ggbiplot)
library(ggplot2)
library(openxlsx)
library(limma)
library(tidyverse)
library(ggrepel)
library(nlme)

merge_df <- readRDS(paste0(working_dir, "Inputs/AthenaOlink_Time_ART_Paired_Cleaned.rds"))
merge_df <- droplevels(merge_df[which(merge_df$ART==1 & !is.na(merge_df$GROUP3)),])
summary(merge_df$GROUP3)

ecvcpers <- droplevels(merge_df[which(merge_df$GROUP3 %in% c("EC_onART", "VCpersistent_onART")),])
ecvcpers$GROUP3 <- relevel(ecvcpers$GROUP3, "VCpersistent_onART")
summary(ecvcpers$GROUP3) 
ecvctrans <- droplevels(merge_df[which(merge_df$GROUP3 %in% c("EC_onART", "VCtransient_onART")),])
ecvctrans$GROUP3 <- relevel(ecvctrans$GROUP3, "VCtransient_onART")
summary(ecvctrans$GROUP3)
```

#### AGE Matching
Source: https://cran.r-project.org/web/packages/MatchIt/vignettes/MatchIt.html
```{r}
library(MatchIt)

## ECVCPERS
# No matching; constructing a pre-match matchit object
m.out0 <- matchit(GROUP3 ~ AGE, data = ecvcpers,
                 method = NULL, distance = "glm")
summary(m.out0)

# plot(m.out0, type = "jitter", interactive = FALSE)
plot(m.out0, type = "density", interactive = FALSE,
     which.xs = ~AGE )

## ECVCTRANS
# No matching; constructing a pre-match matchit object
m.out0 <- matchit(GROUP3 ~ AGE, data = ecvctrans,
                 method = NULL, distance = "glm")
summary(m.out0)

# plot(m.out0, type = "jitter", interactive = FALSE)
plot(m.out0, type = "density", interactive = FALSE,
     which.xs = ~AGE )

```

```{r}
#### ECVCpers
# 1:n Nearest neighbour (NN) propensity score (PS) matching w/o replacement
m.out1 <- matchit(GROUP3 ~ AGE, data = ecvcpers,
                 method = "nearest", distance = "glm", ratio = 2)
summary(m.out1)

plot(m.out1, type = "jitter", interactive = FALSE)
plot(m.out1, type = "density", interactive = FALSE,
     which.xs = ~AGE)

# matched dataset:
ecvcpers_matched <- match.data(m.out1)
summary(ecvcpers_matched$subclass)

### ECVCtrans
# 1:n Nearest neighbour (NN) propensity score (PS) matching w/o replacement
m.out1 <- matchit(GROUP3 ~ AGE, data = ecvctrans,
                 method = "nearest", distance = "glm", ratio = 2)
summary(m.out1)

plot(m.out1, type = "jitter", interactive = FALSE)
plot(m.out1, type = "density", interactive = FALSE,
     which.xs = ~AGE)

# matched dataset:
ecvctrans_matched <- match.data(m.out1)
summary(ecvcpers_matched$subclass)

keep_ids <- unique(c(ecvcpers_matched$ID, ecvctrans_matched$ID))
matched_df <- merge_df[which(merge_df$ID %in% keep_ids),]
summary(matched_df$GROUP3)

# ages table
summary_table <- matched_df %>%
  group_by(GROUP3) %>%
  summarise(val = paste0(round(median(AGE, na.rm = TRUE),1), " (", round(IQR(AGE, na.rm = TRUE),1), ")")) %>%
  pivot_wider(names_from = GROUP3, values_from = val) %>%
  mutate(stat = "median age (IQR)") %>%
  select(stat, everything())

# Display summary_table
print(summary_table)

```


##### Run lmm
```{r}
merge_df <- matched_df

example <- data.frame(NA,NA,NA,NA,NA,NA,NA,NA)
colnames(example) <- c("Outcome","Covariate","Value","Std.Error","DF","t-value","p-value","fdr")
summarystatistics <- list(example,example,example)
names(summarystatistics) <-  c("EC_ART","VCpersistent_ART","VCtransient_ART")

lmm_on_subset <- function(analysisname, treatment, reference, covar_text) {
mergedf1 <- droplevels(merge_df[which(merge_df$GROUP3 == treatment|
                                      merge_df$GROUP3 == reference),])
olinkcolumns <- c((grep("GROUP3",colnames(merge_df))+1): grep("NFU1",colnames(merge_df)))

# run the mixed linear model regression
mod2 <- NULL 
system.time( 
for (i in olinkcolumns) {
  regression_formula <- as.formula(paste(names(mergedf1)[i],"~ GROUP3 +",covar_text)) # make a formula with the olink column name
  lmod <- nlme::lme(regression_formula,
                    random = ~ 1|ID, # the IDs are set as random effect here
                    data = mergedf1,
                    na.action=na.omit,
                    control =list(msMaxIter = 1000, msMaxEval = 1000))
  
  # extract statistics
  stab <- t(summary(lmod)$tTable[2,]) %>% as.data.frame()
  stab$Covariate <-  analysisname
  stab$Outcome <- colnames(merge_df)[i]
  stab %>%
    dplyr::select(Outcome, Covariate, everything()) ->
    stab
  mod2 <- rbind(mod2, stab)
})
mod2$fdr <- p.adjust(mod2$`p-value`, method = 'BH')
mod2 <- mod2[order(mod2$`p-value`),] # order by P value

return (mod2)
}

summarystatistics[["EC_ART"]] <- lmm_on_subset("EC_ART", "EC_onART", "EC_preART", "AGE + STORAGE_TIME")
summarystatistics[["VCpersistent_ART"]] <- lmm_on_subset("VCpersistent_ART", "VCpersistent_onART", "VCpersistent_preART", "AGE + SEX + STORAGE_TIME + CENTER")
summarystatistics[["VCtransient_ART"]] <- lmm_on_subset("VCtransient_ART", "VCtransient_onART", "VCtransient_preART", "AGE + SEX + STORAGE_TIME + CENTER")

summarystatistics

# change list names so it includes sample size: 
names(summarystatistics) <- paste0(names(summarystatistics),"_",c("9","18","18"))
# c("EC_TIME_19","EC_ART_9","VCpersistent_TIME_23","VCpersistent_ART_57","VCtransient_ART_36")

openxlsx::write.xlsx(summarystatistics,paste0(working_dir,"Outputs/RebuttalAddition_AGEMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"))

```

#### RebuttalFigure - volcano VCpersistent
```{r}
# read the summary statistics for VC persistent
result2 <- openxlsx::read.xlsx(paste0(working_dir, "Outputs/RebuttalAddition_AGEMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"), sheet = "VCpersistent_ART_18")

diff_df <- result2[,c("Outcome","p-value","fdr","Value")]
colnames(diff_df) <- c("PROTEIN","RAW_P", "FDR_P","LOG_FC")

# add a grouping column for coloring by significance; default value is "not significant"
diff_df <- diff_df %>% mutate(group= case_when(FDR_P < 0.05 ~ "FDR Significant",
                                               RAW_P < 0.05 & FDR_P > 0.05 ~ "Nominal Significant",
                                               RAW_P > 0.05 ~ "NotSignificant"
                                               ))

# add a text label if 'FDR-P <0.01', or if 'FDR-P <0.05 but LOG_FC >2', or if CRTAM, LY9, CD6, VAT1.
for (i in 1:nrow(diff_df)){
  if(diff_df$RAW_P[i] < 0.001 | # 'FDR-P <0.01'
     (diff_df$RAW_P[i] < 0.01 & abs(diff_df$LOG_FC[i]) > 2 | # 'FDR-P <0.05 but LOG_FC >2'
     (diff_df$PROTEIN[i] %in% c("CRTAM","LY9","CD6","VAT1") # CRTAM, LY9, CD6, VAT1
     )))
  {diff_df$LABEL[i] <- diff_df$PROTEIN[i]} else
  {diff_df$LABEL[i] <- ""}}
diff_df

# make the volcano plot
options(ggrepel.max.overlaps = Inf)
g <- ggplot(diff_df, aes(x=LOG_FC, y=-log10(RAW_P))) +
  geom_point(aes(color=group), alpha=0.8, size=2) +
  scale_color_manual(values=c("orange","#4d4d4d")) + 
  theme_bw() + # size of plotting window
  theme(legend.background=element_rect(),
        plot.title=element_text(angle=0, size=12, face="bold", vjust=1),
        panel.grid.major=element_blank(), #Remove gridlines
        panel.grid.minor=element_blank(), #Remove gridlines
        axis.text.x=element_text(angle=0, size=8, vjust=1),
        axis.text.y=element_text(angle=0, size=8, vjust=1),
        axis.title=element_text(size=10),
        legend.position = "none") +
  # guides(colour = guide_legend(override.aes=list(size=2.5))) + #  Change the size of the icons/symbols in the legend
  xlab("Log2 Fold Change NPX") +
  ylab("-Log10 P value") +
  ggtitle("18 persistent VC paired samples ART" ) +
  geom_label_repel(  #Add the text labels 
    label.size = .25, #Size of label border, in mm.
    box.padding = unit(0.45, "lines"),
    label.padding = 0.25, # Amount of padding around label, as unit or number.
    label.r = 0.15, #Radius of rounded corners,
    size = 3, aes(label = LABEL,hjust = .5))+
  #xlim(-5,3) + ylim(0,1.5) + # limits of the graph
  geom_vline(xintercept=c(0), linetype="longdash", colour="black", size=0.4) + # Add a vertical line for Estimate change cut-offs
  geom_hline(yintercept=-log10(0.05), linetype="longdash", colour="black", size=0.4) #Add a horizontal line for P-value cut-off
g
ggsave(paste0(working_dir, "Outputs/RebuttalAdition_ECVCART_Volcano_label_0.01FDRP_2FC_57VCpersistent_ART_2420_age_storage.pdf"), plot = g, device = NULL, path = NULL, width = 4, height =4, units = 'in', dpi = 300)

# Check direction:
mergedf1 <- droplevels(merge_df[which(merge_df$GROUP3 == "VCpersistent_onART"|
                                      merge_df$GROUP3 == "VCpersistent_preART"),])
boxplot(mergedf1$CD6~as.factor(mergedf1$GROUP3))
boxplot(mergedf1$LY9~as.factor(mergedf1$GROUP3)) 
```

#### RebuttalFigure  - volcano VCtransient
```{r}
# read the summary statistics for VC persistent
result2 <- openxlsx::read.xlsx(paste0(working_dir, "Outputs/RebuttalAddition_AGEMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"), sheet = "VCtransient_ART_18")

diff_df <- result2[,c("Outcome","p-value","fdr","Value")]
colnames(diff_df) <- c("PROTEIN","RAW_P", "FDR_P","LOG_FC")

# add a grouping column for coloring by significance; default value is "not significant"
diff_df <- diff_df %>% mutate(group= case_when(FDR_P < 0.05 ~ "FDR Significant",
                                               RAW_P < 0.05 & FDR_P > 0.05 ~ "Nominal Significant",
                                               RAW_P > 0.05 ~ "NotSignificant"
                                               ))

# add a text label if 'FDR-P <0.05', or if 'FDR-P <0.05 but LOG_FC >2', or if CRTAM, LY9, CD6, VAT1.
for (i in 1:nrow(diff_df)){
  if(diff_df$RAW_P[i] < 0.001 | # 'FDR-P <0.05'
     (diff_df$RAW_P[i] < 0.01 & abs(diff_df$LOG_FC[i]) > 2 | # 'FDR-P <0.05 but LOG_FC >2'
     (diff_df$PROTEIN[i] %in% c("CRTAM","LY9","CD6","VAT1") # CRTAM, LY9, CD6, VAT1
     )))
  {diff_df$LABEL[i] <- diff_df$PROTEIN[i]} else
  {diff_df$LABEL[i] <- ""}}
diff_df

# make the volcano plot
options(ggrepel.max.overlaps = Inf)
g <- ggplot(diff_df, aes(x=LOG_FC, y=-log10(RAW_P))) +
  geom_point(aes(color=group), alpha=0.8, size=2) +
  scale_color_manual(values=c("orange","#4d4d4d")) + 
  theme_bw() + # size of plotting window
  theme(legend.background=element_rect(),
        plot.title=element_text(angle=0, size=12, face="bold", vjust=1),
        panel.grid.major=element_blank(), #Remove gridlines
        panel.grid.minor=element_blank(), #Remove gridlines
        axis.text.x=element_text(angle=0, size=8, vjust=1),
        axis.text.y=element_text(angle=0, size=8, vjust=1),
        axis.title=element_text(size=10),
        legend.position = "none") +
  # guides(colour = guide_legend(override.aes=list(size=2.5))) + #  Change the size of the icons/symbols in the legend
  xlab("Log2 Fold Change NPX") +
  ylab("-Log10 P value") +
  ggtitle("18 transient VC paired samples ART" ) +
  geom_label_repel(  #Add the text labels 
    label.size = .25, #Size of label border, in mm.
    box.padding = unit(0.45, "lines"),
    label.padding = 0.25, # Amount of padding around label, as unit or number.
    label.r = 0.15, #Radius of rounded corners,
    size = 2.75, aes(label = LABEL,hjust = .5))+
  #xlim(-4.1,3) + ylim(0, 6) + # limits of the graph
  geom_vline(xintercept=c(0), linetype="longdash", colour="black", size=0.4) #+ # Add a vertical line for Estimate change cut-offs
  #geom_hline(yintercept= log10(0.05), linetype="longdash", colour="black", size=0.4) #Add a horizontal line for P-value cut-off
g

ggsave(paste0(working_dir, "Outputs/RebuttalAdditionECVCART_Volcano_label_0.001FDRP_2FC_36VCtransient_ART_2420_age_storage.pdf"), plot = g, width = 4, height =4, units = 'in', dpi = 300)
```

#### RebuttalFigure - volcano ECpersistent
```{r}
# read the summary statistics for VC persistent
result2 <- openxlsx::read.xlsx(paste0(working_dir, "Outputs/RebuttalAddition_AGEMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"), sheet = "EC_ART_9")

diff_df <- result2[,c("Outcome","p-value","fdr","Value")]
colnames(diff_df) <- c("PROTEIN","RAW_P", "FDR_P","LOG_FC")

# add a grouping column for coloring by significance; default value is "not significant"
diff_df <- diff_df %>% mutate(group= case_when(FDR_P < 0.05 ~ "FDR Significant",
                                               RAW_P < 0.05 & FDR_P > 0.05 ~ "Nominal Significant",
                                               RAW_P > 0.05 ~ "NotSignificant"
                                               ))

# add a text label if 'FDR-P <0.05', or if 'FDR-P <0.05 but LOG_FC >2', or if CRTAM, LY9, CD6, VAT1.
for (i in 1:nrow(diff_df)){
  if(diff_df$RAW_P[i] < 0.01 | # 'FDR-P <0.05'
     (diff_df$FDR_P[i] < 0.05 & abs(diff_df$LOG_FC[i]) > 2 | # 'FDR-P <0.05 but LOG_FC >2'
     (diff_df$PROTEIN[i] %in% c("CRTAM","LY9","CD6","VAT1") # CRTAM, LY9, CD6, VAT1
     )))
  {diff_df$LABEL[i] <- diff_df$PROTEIN[i]} else
  {diff_df$LABEL[i] <- ""}}
diff_df

# make the volcano plot
options(ggrepel.max.overlaps = Inf)
g <- ggplot(diff_df, aes(x=LOG_FC, y=-log10(RAW_P))) +
  geom_point(aes(color=group), alpha=0.8, size=2) +
  scale_color_manual(values=c("orange","#4d4d4d")) + # no red because no FDR significant
  theme_bw() + # size of plotting window
  theme(legend.background=element_rect(),
        plot.title=element_text(angle=0, size=12, face="bold", vjust=1),
        panel.grid.major=element_blank(), #Remove gridlines
        panel.grid.minor=element_blank(), #Remove gridlines
        axis.text.x=element_text(angle=0, size=8, vjust=1),
        axis.text.y=element_text(angle=0, size=8, vjust=1),
        axis.title=element_text(size=10),
        legend.position = "none") +
  # guides(colour = guide_legend(override.aes=list(size=2.5))) + #  Change the size of the icons/symbols in the legend
  xlab("Log2 Fold Change NPX") +
  ylab("-Log10 P value") +
  ggtitle("9 persistent EC paired samples ART" ) +
  geom_label_repel(  #Add the text labels 
    label.size = .25, #Size of label border, in mm.
    box.padding = unit(0.45, "lines"),
    label.padding = 0.25, # Amount of padding around label, as unit or number.
    label.r = 0.15, #Radius of rounded corners,
    size = 2.75, aes(label = LABEL,hjust = .5))+
  xlim(-4.1,3.4) + ylim(0, 4.5) + # limits of the graph
  geom_vline(xintercept=c(0), linetype="longdash", colour="black", size=0.4) + # Add a vertical line for Estimate change cut-offs
  geom_hline(yintercept= -log10(0.05), linetype="longdash", colour="black", size=0.4) #Add a horizontal line for P-value cut-off
g

ggsave(paste0(working_dir, "Outputs/RebuttalAddtion_ECVCART_Volcano_label_0.001RAWP_2FC_9EC_ART_2420_age_storage.pdf"), plot = g, device = NULL, path = NULL,scale = 1, width = 4, height =4, units = 'in', dpi = 300)
```

####
####
####
### ART-duration MATCHED comparisons
```{r}
merge_df <- readRDS(paste0(working_dir, "Inputs/AthenaOlink_Time_ART_Paired_Cleaned.rds"))
merge_df <- droplevels(merge_df[which(merge_df$ART==1 & !is.na(merge_df$GROUP3)),])
summary(merge_df$GROUP3)

ecvcpers <- droplevels(merge_df[which(merge_df$GROUP3 %in% c("EC_onART", "VCpersistent_onART")),])
ecvcpers$GROUP3 <- relevel(ecvcpers$GROUP3, "VCpersistent_onART")
summary(ecvcpers$GROUP3) 
ecvctrans <- droplevels(merge_df[which(merge_df$GROUP3 %in% c("EC_onART", "VCtransient_onART")),])
ecvctrans$GROUP3 <- relevel(ecvctrans$GROUP3, "VCtransient_onART")
summary(ecvctrans$GROUP3)
summary(merge_df$T2_ART_DUR_DAYS)
```
#### AGE Matching
Source: https://cran.r-project.org/web/packages/MatchIt/vignettes/MatchIt.html
```{r}
library(MatchIt)
merge_df
## ECVCPERS
# No matching; constructing a pre-match matchit object
m.out0 <- matchit(GROUP3 ~ T2_ART_DUR_DAYS, data = ecvcpers,
                 method = NULL, distance = "glm")
summary(m.out0)

# plot(m.out0, type = "jitter", interactive = FALSE)
plot(m.out0, type = "density", interactive = FALSE,
     which.xs = ~T2_ART_DUR_DAYS )

## ECVCTRANS
# No matching; constructing a pre-match matchit object
m.out0 <- matchit(GROUP3 ~ T2_ART_DUR_DAYS, data = ecvctrans,
                 method = NULL, distance = "glm")
summary(m.out0)

# plot(m.out0, type = "jitter", interactive = FALSE)
plot(m.out0, type = "density", interactive = FALSE,
     which.xs = ~T2_ART_DUR_DAYS )

```

```{r}
#### ECVCpers
# 1:n Nearest neighbour (NN) propensity score (PS) matching w/o replacement
m.out1 <- matchit(GROUP3 ~ T2_ART_DUR_DAYS, data = ecvcpers,
                 method = "nearest", distance = "glm", ratio = 2)
summary(m.out1)

plot(m.out1, type = "jitter", interactive = FALSE)
plot(m.out1, type = "density", interactive = FALSE,
     which.xs = ~T2_ART_DUR_DAYS)

# matched dataset:
ecvcpers_matched <- match.data(m.out1)
summary(ecvcpers_matched$subclass)

### ECVCtrans
# 1:n Nearest neighbour (NN) propensity score (PS) matching w/o replacement
m.out1 <- matchit(GROUP3 ~ T2_ART_DUR_DAYS, data = ecvctrans,
                 method = "nearest", distance = "glm", ratio = 2)
summary(m.out1)

plot(m.out1, type = "jitter", interactive = FALSE)
plot(m.out1, type = "density", interactive = FALSE,
     which.xs = ~T2_ART_DUR_DAYS)

# matched dataset:
ecvctrans_matched <- match.data(m.out1)
summary(ecvcpers_matched$subclass)

keep_ids <- unique(c(ecvcpers_matched$ID, ecvctrans_matched$ID))
matched_df <- merge_df[which(merge_df$ID %in% keep_ids),]
summary(matched_df$GROUP3)

# ages table
summary_table <- matched_df %>%
  group_by(GROUP3) %>%
  summarise(val = paste0(round(median(T2_ART_DUR_DAYS/365.25, na.rm = TRUE),1), " (", round(IQR(T2_ART_DUR_DAYS/365.25, na.rm = TRUE),1), ")")) %>%
  pivot_wider(names_from = GROUP3, values_from = val) %>%
  mutate(stat = "median ART duration in years (IQR)") %>%
  select(stat, everything())

# Display summary_table
print(summary_table)

```
##### Run final model
```{r}
merge_df <- matched_df

example <- data.frame(NA,NA,NA,NA,NA,NA,NA,NA)
colnames(example) <- c("Outcome","Covariate","Value","Std.Error","DF","t-value","p-value","fdr")
summarystatistics <- list(example,example,example)
names(summarystatistics) <-  c("EC_ART","VCpersistent_ART","VCtransient_ART")

lmm_on_subset <- function(analysisname, treatment, reference, covar_text) {
mergedf1 <- droplevels(merge_df[which(merge_df$GROUP3 == treatment|
                                      merge_df$GROUP3 == reference),])
olinkcolumns <- c((grep("GROUP3",colnames(merge_df))+1): grep("NFU1",colnames(merge_df)))

# run the mixed linear model regression
mod2 <- NULL 
system.time( 
for (i in olinkcolumns) {
  regression_formula <- as.formula(paste(names(mergedf1)[i],"~ GROUP3 +",covar_text)) # make a formula with the olink column name
  lmod <- nlme::lme(regression_formula,
                    random = ~ 1|ID, # the IDs are set as random effect here
                    data = mergedf1,
                    na.action=na.omit,
                    control =list(msMaxIter = 1000, msMaxEval = 1000))
  
  # extract statistics
  stab <- t(summary(lmod)$tTable[2,]) %>% as.data.frame()
  stab$Covariate <-  analysisname
  stab$Outcome <- colnames(merge_df)[i]
  stab %>%
    dplyr::select(Outcome, Covariate, everything()) ->
    stab
  mod2 <- rbind(mod2, stab)
})
mod2$fdr <- p.adjust(mod2$`p-value`, method = 'BH')
mod2 <- mod2[order(mod2$`p-value`),] # order by P value

return (mod2)
}

summarystatistics[["EC_ART"]] <- lmm_on_subset("EC_ART", "EC_onART", "EC_preART", "AGE + STORAGE_TIME")
summarystatistics[["VCpersistent_ART"]] <- lmm_on_subset("VCpersistent_ART", "VCpersistent_onART", "VCpersistent_preART", "AGE + SEX + STORAGE_TIME + CENTER")
summarystatistics[["VCtransient_ART"]] <- lmm_on_subset("VCtransient_ART", "VCtransient_onART", "VCtransient_preART", "AGE + SEX + STORAGE_TIME + CENTER")

summarystatistics

# change list names so it includes sample size: 
names(summarystatistics) <- paste0(names(summarystatistics),"_",c("9","18","18"))
# c("EC_TIME_19","EC_ART_9","VCpersistent_TIME_23","VCpersistent_ART_57","VCtransient_ART_36")

openxlsx::write.xlsx(summarystatistics,paste0(working_dir,"Outputs/RebuttalAddition_ARTDURMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"))

```

#### RebuttalFigure - volcano VCpersistent
```{r}
# read the summary statistics for VC persistent
result2 <- openxlsx::read.xlsx(paste0(working_dir, "Outputs/RebuttalAddition_ARTDURMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"), sheet = "VCpersistent_ART_18")

diff_df <- result2[,c("Outcome","p-value","fdr","Value")]
colnames(diff_df) <- c("PROTEIN","RAW_P", "FDR_P","LOG_FC")

# add a grouping column for coloring by significance; default value is "not significant"
diff_df <- diff_df %>% mutate(group= case_when(FDR_P < 0.05 ~ "FDR Significant",
                                               RAW_P < 0.05 & FDR_P > 0.05 ~ "Nominal Significant",
                                               RAW_P > 0.05 ~ "NotSignificant"
                                               ))

# add a text label if 'FDR-P <0.01', or if 'FDR-P <0.05 but LOG_FC >2', or if CRTAM, LY9, CD6, VAT1.
for (i in 1:nrow(diff_df)){
  if(diff_df$RAW_P[i] < 0.001 | # 'FDR-P <0.01'
     (diff_df$RAW_P[i] < 0.01 & abs(diff_df$LOG_FC[i]) > 2 | # 'FDR-P <0.05 but LOG_FC >2'
     (diff_df$PROTEIN[i] %in% c("CRTAM","LY9","CD6","VAT1") # CRTAM, LY9, CD6, VAT1
     )))
  {diff_df$LABEL[i] <- diff_df$PROTEIN[i]} else
  {diff_df$LABEL[i] <- ""}}
diff_df

# make the volcano plot
options(ggrepel.max.overlaps = Inf)
g <- ggplot(diff_df, aes(x=LOG_FC, y=-log10(RAW_P))) +
  geom_point(aes(color=group), alpha=0.8, size=2) +
  scale_color_manual(values=c("orange","#4d4d4d")) + 
  theme_bw() + # size of plotting window
  theme(legend.background=element_rect(),
        plot.title=element_text(angle=0, size=12, face="bold", vjust=1),
        panel.grid.major=element_blank(), #Remove gridlines
        panel.grid.minor=element_blank(), #Remove gridlines
        axis.text.x=element_text(angle=0, size=8, vjust=1),
        axis.text.y=element_text(angle=0, size=8, vjust=1),
        axis.title=element_text(size=10),
        legend.position = "none") +
  # guides(colour = guide_legend(override.aes=list(size=2.5))) + #  Change the size of the icons/symbols in the legend
  xlab("Log2 Fold Change NPX") +
  ylab("-Log10 P value") +
  ggtitle("18 persistent VC paired samples ART" ) +
  geom_label_repel(  #Add the text labels 
    label.size = .25, #Size of label border, in mm.
    box.padding = unit(0.45, "lines"),
    label.padding = 0.25, # Amount of padding around label, as unit or number.
    label.r = 0.15, #Radius of rounded corners,
    size = 3, aes(label = LABEL,hjust = .5))+
  #xlim(-5,3) + ylim(0,1.5) + # limits of the graph
  geom_vline(xintercept=c(0), linetype="longdash", colour="black", size=0.4) + # Add a vertical line for Estimate change cut-offs
  geom_hline(yintercept=-log10(0.05), linetype="longdash", colour="black", size=0.4) #Add a horizontal line for P-value cut-off
g
ggsave(paste0(working_dir, "Outputs/RebuttalAdditionARTDUR_ECVCART_Volcano_label_0.01FDRP_2FC_57VCpersistent_ART_2420_age_storage.pdf"), plot = g, device = NULL, path = NULL, width = 4, height =4, units = 'in', dpi = 300)

# Check direction:
mergedf1 <- droplevels(merge_df[which(merge_df$GROUP3 == "VCpersistent_onART"|
                                      merge_df$GROUP3 == "VCpersistent_preART"),])
boxplot(mergedf1$CD6~as.factor(mergedf1$GROUP3))
boxplot(mergedf1$LY9~as.factor(mergedf1$GROUP3)) 
```

#### RebuttalFigure  - volcano VCtransient
```{r}
# read the summary statistics for VC persistent
result2 <- openxlsx::read.xlsx(paste0(working_dir, "Outputs/RebuttalAddition_ARTDURMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"), sheet = "VCtransient_ART_18")

diff_df <- result2[,c("Outcome","p-value","fdr","Value")]
colnames(diff_df) <- c("PROTEIN","RAW_P", "FDR_P","LOG_FC")

# add a grouping column for coloring by significance; default value is "not significant"
diff_df <- diff_df %>% mutate(group= case_when(FDR_P < 0.05 ~ "FDR Significant",
                                               RAW_P < 0.05 & FDR_P > 0.05 ~ "Nominal Significant",
                                               RAW_P > 0.05 ~ "NotSignificant"
                                               ))

# add a text label if 'FDR-P <0.05', or if 'FDR-P <0.05 but LOG_FC >2', or if CRTAM, LY9, CD6, VAT1.
for (i in 1:nrow(diff_df)){
  if(diff_df$RAW_P[i] < 0.001 | # 'FDR-P <0.05'
     (diff_df$RAW_P[i] < 0.01 & abs(diff_df$LOG_FC[i]) > 2 | # 'FDR-P <0.05 but LOG_FC >2'
     (diff_df$PROTEIN[i] %in% c("CRTAM","LY9","CD6","VAT1") # CRTAM, LY9, CD6, VAT1
     )))
  {diff_df$LABEL[i] <- diff_df$PROTEIN[i]} else
  {diff_df$LABEL[i] <- ""}}
diff_df

# make the volcano plot
options(ggrepel.max.overlaps = Inf)
g <- ggplot(diff_df, aes(x=LOG_FC, y=-log10(RAW_P))) +
  geom_point(aes(color=group), alpha=0.8, size=2) +
  scale_color_manual(values=c("#b2182b","orange","#4d4d4d")) + 
  theme_bw() + # size of plotting window
  theme(legend.background=element_rect(),
        plot.title=element_text(angle=0, size=12, face="bold", vjust=1),
        panel.grid.major=element_blank(), #Remove gridlines
        panel.grid.minor=element_blank(), #Remove gridlines
        axis.text.x=element_text(angle=0, size=8, vjust=1),
        axis.text.y=element_text(angle=0, size=8, vjust=1),
        axis.title=element_text(size=10),
        legend.position = "none") +
  # guides(colour = guide_legend(override.aes=list(size=2.5))) + #  Change the size of the icons/symbols in the legend
  xlab("Log2 Fold Change NPX") +
  ylab("-Log10 P value") +
  ggtitle("18 transient VC paired samples ART" ) +
  geom_label_repel(  #Add the text labels 
    label.size = .25, #Size of label border, in mm.
    box.padding = unit(0.45, "lines"),
    label.padding = 0.25, # Amount of padding around label, as unit or number.
    label.r = 0.15, #Radius of rounded corners,
    size = 2.75, aes(label = LABEL,hjust = .5))+
  #xlim(-4.1,3) + ylim(0, 6) + # limits of the graph
  geom_vline(xintercept=c(0), linetype="longdash", colour="black", size=0.4) #+ # Add a vertical line for Estimate change cut-offs
  #geom_hline(yintercept= log10(0.05), linetype="longdash", colour="black", size=0.4) #Add a horizontal line for P-value cut-off
g

ggsave(paste0(working_dir, "Outputs/RebuttalAdditionARTDUR_ECVCART_Volcano_label_0.001FDRP_2FC_36VCtransient_ART_2420_age_storage.pdf"), plot = g, width = 4, height =4, units = 'in', dpi = 300)
```

#### RebuttalFigure - volcano ECpersistent
```{r}
# read the summary statistics for VC persistent
result2 <- openxlsx::read.xlsx(paste0(working_dir, "Outputs/RebuttalAddition_ARTDURMATCHED_LMM_Olink_ECVC_ARTandTIME_2420_age_storage_sex_center.xlsx"), sheet = "EC_ART_9")

diff_df <- result2[,c("Outcome","p-value","fdr","Value")]
colnames(diff_df) <- c("PROTEIN","RAW_P", "FDR_P","LOG_FC")

# add a grouping column for coloring by significance; default value is "not significant"
diff_df <- diff_df %>% mutate(group= case_when(FDR_P < 0.05 ~ "FDR Significant",
                                               RAW_P < 0.05 & FDR_P > 0.05 ~ "Nominal Significant",
                                               RAW_P > 0.05 ~ "NotSignificant"
                                               ))

# add a text label if 'FDR-P <0.05', or if 'FDR-P <0.05 but LOG_FC >2', or if CRTAM, LY9, CD6, VAT1.
for (i in 1:nrow(diff_df)){
  if(diff_df$RAW_P[i] < 0.01 | # 'FDR-P <0.05'
     (diff_df$FDR_P[i] < 0.05 & abs(diff_df$LOG_FC[i]) > 2 | # 'FDR-P <0.05 but LOG_FC >2'
     (diff_df$PROTEIN[i] %in% c("CRTAM","LY9","CD6","VAT1") # CRTAM, LY9, CD6, VAT1
     )))
  {diff_df$LABEL[i] <- diff_df$PROTEIN[i]} else
  {diff_df$LABEL[i] <- ""}}
diff_df

# make the volcano plot
options(ggrepel.max.overlaps = Inf)
g <- ggplot(diff_df, aes(x=LOG_FC, y=-log10(RAW_P))) +
  geom_point(aes(color=group), alpha=0.8, size=2) +
  scale_color_manual(values=c("orange","#4d4d4d")) + # no red because no FDR significant
  theme_bw() + # size of plotting window
  theme(legend.background=element_rect(),
        plot.title=element_text(angle=0, size=12, face="bold", vjust=1),
        panel.grid.major=element_blank(), #Remove gridlines
        panel.grid.minor=element_blank(), #Remove gridlines
        axis.text.x=element_text(angle=0, size=8, vjust=1),
        axis.text.y=element_text(angle=0, size=8, vjust=1),
        axis.title=element_text(size=10),
        legend.position = "none") +
  # guides(colour = guide_legend(override.aes=list(size=2.5))) + #  Change the size of the icons/symbols in the legend
  xlab("Log2 Fold Change NPX") +
  ylab("-Log10 P value") +
  ggtitle("9 persistent EC paired samples ART" ) +
  geom_label_repel(  #Add the text labels 
    label.size = .25, #Size of label border, in mm.
    box.padding = unit(0.45, "lines"),
    label.padding = 0.25, # Amount of padding around label, as unit or number.
    label.r = 0.15, #Radius of rounded corners,
    size = 2.75, aes(label = LABEL,hjust = .5))+
  xlim(-4.1,3.4) + ylim(0, 4.5) + # limits of the graph
  geom_vline(xintercept=c(0), linetype="longdash", colour="black", size=0.4) + # Add a vertical line for Estimate change cut-offs
  geom_hline(yintercept= -log10(0.05), linetype="longdash", colour="black", size=0.4) #Add a horizontal line for P-value cut-off
g

ggsave(paste0(working_dir, "Outputs/RebuttalAdditionARTDUR_ECVCART_Volcano_label_0.001RAWP_2FC_9EC_ART_2420_age_storage.pdf"), plot = g, device = NULL, path = NULL,scale = 1, width = 4, height =4, units = 'in', dpi = 300)
```






