---
title: "Olink ananlysis eliteATHENA and 2000HIV"
author: "Albert Groenendijk & Nadira Vadaq"
date: "2022-11-22"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    number_sections: yes
---

The third reviewer asked about the possibility that differencees are only present for HIV controllers with higher BMI >25. This is an in-depth analysis of the role of BMI in our data.

```{r echo=T, results='hide', warning=F,message=F}
working_dir <- c("HIV_controllers_ATHENA/AthenaOlink_DifferentialExpression_CrossSectional/")

library(FactoMineR)
library(dplyr)
library(tidyr)
library(magrittr)
library(lubridate)
library(table1)
library(ggbiplot)
library(ggplot2)
library(gridExtra)
library(openxlsx)
```


## Read Cleaned Data AthenaOlink

```{r, warning=FALSE}
master_df = readRDS(paste0(working_dir, "Inputs/AthenaOlink_EC_vs_VC_Cleaned.rds"))
```

## Visualize BMI distribution
```{r}
summary(master_df$T1_BMI)
summary(master_df$BMI_CAT)

master_df %>%
  group_by(NONVIREMIC) %>%
  summarize(
    missing_count = sum(is.na(T1_BMI)),
    nonmissing_count = sum(!is.na(T1_BMI)),
    total = n(),
    percentage_missing = round((missing_count / total) * 100,2)
  )

ggplot(master_df, aes(x = T1_BMI, color = NONVIREMIC)) +
  geom_density(linewidth= 2) +
  labs(
    title = "Density Distribution of BMI by NONVIREMIC Status",
    x = "BMI",
    y = "Density",
    color = "NONVIREMIC"
  ) +
  theme_minimal()
```

## counts of BMI CAT2 per stratum
```{r}
cocktails_df <- master_df %>%
  dplyr::select( "ID", "NONVIREMIC","SMOKING_CURRENT","T1_BMI","VL",
                 "CRTAM_Neurology","LY9_Inflammation","CD6_Inflammation","VAT1_Oncology") %>%
  mutate(BMI_cat2 = case_when(T1_BMI <18.5 ~ "BMI <18.5 (N)",
                             T1_BMI >=18.5 & T1_BMI <25 ~ "BMI 18.5-24.9 (N)",
                             T1_BMI >=25 ~ "BMI >=25 (N)"))

cocktails_df %>%
  filter(!is.na(BMI_cat2)) %>%   # exclude missing BMI categories
  count(NONVIREMIC, BMI_cat2) %>%
  pivot_wider(names_from = BMI_cat2, values_from = n, values_fill = 0) %>%
  select(NONVIREMIC, `BMI <18.5 (N)`, `BMI 18.5-24.9 (N)`, `BMI >=25 (N)`) %>%
  rename(Group = NONVIREMIC)
```
## Correlate BMI with VL/AGE/CRTAM/LY9/CD6/VAT1
```{r}
df_corr <- master_df %>%
  select(
    ID = COMP_1_2_ID,
    BMI = T1_BMI,
    VL = VL,
    AGE = AGE,
    CRTAM = CRTAM_Neurology,
    LY9 = LY9_Inflammation,
    CD6 = CD6_Inflammation,
    VAT1 = VAT1_Oncology,
    ECVC = NONVIREMIC
  ) %>% filter(!is.na(BMI))

y_cols <- c("VL", "AGE", "CRTAM", "LY9","CD6", "VAT1")

plots <- list()
for (y_col in y_cols) {
  # Get correlation test results between BMI and y_col
  corr_test <- cor.test(df_corr$BMI, df_corr[[y_col]], use = "complete.obs", method='spearman')
  r_val <- round(corr_test$estimate, 2)
  p_val <- signif(corr_test$p.value, 2)
  
  # Create subtitle with correlation and missing BMI count info
  subtitle_text <- paste0("rho = ", r_val, ", p = ", p_val)
                          
  
  # Plot with ggplot
  p <- ggplot(df_corr, aes(x = .data$BMI, y = .data[[y_col]])) +
    geom_point(alpha = 0.6, size=0.5) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = paste("BMI against", y_col),
      subtitle = subtitle_text,
      x = "BMI (kg/m²)",
      y = y_col
    ) +
    theme_minimal()
  
  plots[[y_col]] <- p
  
}
# Arrange plots in 3 columns and 2 rows
combined_plot <- grid.arrange(grobs = plots, ncol = 3, nrow = 2)

# Save to PDF: dimensions of A4 size in inches ~ 8.27 x 11.69
# Set width = 8.27 (A4 width), height = 11.69 / 3 ~ 3.9 (one third A4 height)
ggsave(paste0(working_dir, "Outputs/AdditionalRebuttal_BMICorrelations.pdf"), combined_plot, dpi=300, width = 8.27, height = 5, units = "in")
```
#### prepare data
```{r}
library("reshape2")

# Prepare the data
cocktails_df <- master_df %>%
  dplyr::select( "ID", "NONVIREMIC","SMOKING_CURRENT","T1_BMI","VL",
                 "CRTAM_Neurology","LY9_Inflammation","CD6_Inflammation","VAT1_Oncology") %>%
  mutate(BMI_cat2 = case_when(T1_BMI <18.5 ~ "BMI <18.5",
                             T1_BMI >=18.5 & T1_BMI <24.9 ~ "BMI 18.5-24.9",
                             T1_BMI >=25 & T1_BMI <30 ~ "BMI 25-29.9",
                             T1_BMI >=30 ~ "BMI >=30"),
         SMOKING_CURRENT = ifelse(SMOKING_CURRENT==0,"Smoking (+)","Smoking (-)"))%>%
  dplyr::mutate(VL_cat = case_when( NONVIREMIC=="EC" ~ "EC",
                            NONVIREMIC=="VC" & VL < 400 ~ "50-400",
                            NONVIREMIC=="VC" & VL >= 400 & VL < 2000 ~ "400-2000",
                            NONVIREMIC=="VC" & VL > 2000 ~ "2000-10000")) #choose cell populations of interest

# pivot longer with melt() from reshape2
melt_data <- reshape2::melt(cocktails_df, id=c("ID", "NONVIREMIC","SMOKING_CURRENT","T1_BMI","BMI_cat2","VL","VL_cat")) 
summary(as.factor(melt_data$variable))
# extract protein and panel names from variable name
melt_data <- melt_data %>%
  mutate(Protein = stringr::str_split_i(variable,"_",1),
         Panel = stringr::str_split_i(variable,"_",2))

# make factors in the right order
melt_data$VL_cat <- factor(melt_data$VL_cat, levels = c("EC","50-400","400-2000","2000-10000"))
melt_data$Protein <- factor(melt_data$Protein, levels = c("CRTAM","LY9","CD6","VAT1"))
melt_data$BMI_cat <- factor(melt_data$BMI_cat, levels = c("BMI <18.5","BMI 18.5-24.9","BMI 25-29.9", "BMI >=30" ))

# double check whether the data has been correctly transformed
# head(melt_data)
# summary(as.factor(melt_data$Protein))
# str(melt_data)
```

#### Redo Supplementary boxplots, stratification by BMI
```{r}
### Figure S2B - BMI stratification
melt_data$BMI_cat2 <- factor(melt_data$BMI_cat2, levels = unique(melt_data$BMI_cat2)[c(5,1,2,4)])

melt_data %>% 
  filter(!is.na(BMI_cat2))%>%
  ggplot(aes(x = BMI_cat2, y = value, fill = NONVIREMIC))+
  #geom_violin(alpha=0.3, position = position_dodge(width = .9), size=.1, color=NA) +
  geom_boxplot(notch = FALSE,  outlier.size = -1, width=.15, 
               #color="black",
               lwd=.5, position = position_dodge(width = .9),
               alpha = 0.5,show.legend = F)+
  geom_jitter( shape= 21, size=.5, 
               position = position_jitterdodge(dodge.width = .9), 
               aes(color = NONVIREMIC),alpha=.5)+
  ylab(c("Log2 NPX"))  +
  xlab(c(""))  +
  facet_wrap(~Protein, scales = "free",ncol = 2)+#
  theme_bw(base_size = 8)+
  guides(fill = guide_legend(override.aes = list(alpha = .5,color="black")))+
  scale_fill_manual(values=c("red3","darkgrey"))+
  scale_color_manual(values=c( "red3","darkgrey"))+
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8, angle = 45, hjust = 1))

ggsave(paste0(working_dir, "Outputs/AdditionalRebuttal_FigureS2B_Boxplot_Stratified_BMIcategory.pdf"), dpi = 300,width = 10, height = 10, units = "cm" ) 

# P values
for(protein in levels(melt_data$Protein)) {
  for(stratum in levels(melt_data$BMI_cat2)) {
  pval <- wilcox.test(melt_data[which(melt_data$Protein == protein &
                                melt_data$BMI_cat2 == stratum),"value"] ~ # by EC/VC status
              melt_data[which(melt_data$Protein == protein &
                                melt_data$BMI_cat2 == stratum),"NONVIREMIC"])$p.value
  print(paste(protein, stratum, pval))
}}
```
#### who are the three outliers?
Anonymized
```{r}
# cocktails_df %>% filter(BMI_cat2 == "BMI 18.5-24.9" & NONVIREMIC=="EC") %>% select(ID, CRTAM_Neurology, LY9_Inflammation)
```




