---
title: "2410246_ClinicalModels"
author: "Albert Groenendijk"
date: "2024-10-26"
output: html_document
---

limma 3.58.1; survival 3.5.8; survminer 0.4.9; slider 0.3.1; RiskRegression 2023.12.21; randomForestSRC 3.3.1; glmnet 4.1.8; gbm 2.2.2


Load libraries
```{r}
library(tidyverse)
library(survival)
library(ggplot2)
library(riskRegression) # for time-dependent AUC calculation
library(randomForestSRC)
library(glmnet) # also loaded as part of riskRegression when you call GLMnet()
library(hal9001) # also loaded as part of riskRegression when you call Hal90001()
```

Define directories
```{r}
inputdir <- "HIV_controllers_ATHENA/AthenaOlinkClinicalCombined_ClinicalPredictionModels/Input/"
scriptsdir <- "HIV_controllers_ATHENA/AthenaOlinkClinicalCombined_ClinicalPredictionModels/Scripts/"
outputdir <- "HIV_controllers_ATHENA/AthenaOlinkClinicalCombined_ClinicalPredictionModels/Output/"
```

Define generic objects
```{r}
`%notin%` <- Negate(`%in%`)
```

Set random seed
```{r}
set.seed(101)
```

Load datasets
```{r}
train <- readRDS(paste0(outputdir,"clinical_train.rds"))
test <- readRDS(paste0(outputdir,"clinical_test.rds"))
```

Verify data integrity
```{r}
# Are the column names equal
all(names(train) == names(test))

# Ensure consistent factor levels in both train and test datasets
factor_vars <- c("MALESEX", "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER", "baseHCVever", "baseHBVever")

## Convert these variables to factors in both datasets and align levels
for (var in factor_vars) {
  train[[var]] <- factor(train[[var]])
  test[[var]] <- factor(test[[var]], levels = levels(train[[var]]))  # Align levels with training data
}

check_data <- FALSE
if (check_data){
  # Check the data structure for both datasets
  str(train)
  str(test)
  
  # Identify missingness
  na_counts <- sapply(train, function(x) sum(is.na(x)))
  print(na_counts)
  
  na_counts <- sapply(test, function(x) sum(is.na(x)))
  print(na_counts)
}
```

# Run univariate Cox regression on train dataset

```{r}
# First modify AGE and BMI, for which one unit is not so relevant.
## Instead, divide by x so that the HR for a change of x units is determined.
train <- train %>%
  mutate(AGE = AGE / 10,
         baseBMI = baseBMI / 5,
         baseCD4_sqrt = baseCD4_sqrt / 5)
    
## Leave baseCD4_sqrt as is since this is strongly associated with CD4_LOSS

# Define features (predictors) and outcomes
features <- c("MALESEX", "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER","baseVL_log10","baseHCV","baseHBV","baseHCVever","baseHBVever","AGE","baseBMI","baseCD4_sqrt")
outcomes <- c("VL4K","HYBRID", "CD4")

# Frequency of outcome happening
for (outcome in outcomes){
  event <- paste0("EVENT_", outcome)
  print(summary.factor(train[,event]))
}

# Define function to run and clean a univariate cox model for a specific feature and outcome
coxph_singlerow <- function(feature, outcome, dataset){
  # Define the formula for the cox model
  time <- paste0("TIME_", outcome)
  event <- paste0("EVENT_", outcome)
  formula_coxph <- as.formula(paste("Surv(", time, ",", event, ") ~", feature))
  
  # filter out rows without outcome, only applicable for EVENT_CD4
  dataset <- train %>% filter(!is.na(.data[[event]]))
  
  # run the univariate model
  model <- coxph(formula_coxph, data = dataset)

  # clean the cox ph model into a tibble 
  ## With columns: estimate, std.error, statistic, p.value, conf.low, conf.high
  sumstats <- broom::tidy(model, conf.int = TRUE, exponentiate = TRUE)

  new_row <- tibble(
    Variable = feature,
    Number = sum(!is.na(dataset[[feature]])),
    Outcome = event,
    Analysis = "univariate",
    HR = sumstats$estimate,
    Lower = sumstats$conf.low,
    Upper = sumstats$conf.high,
    Pvalue = sumstats$p.value
  )
  
  return(new_row)
}

# Example
coxph_singlerow(feature = features[1], outcome = outcomes[1], dataset = dataset)

# map the coxph_singlerow to every feature and outcome
sumstats_univariate_train <- expand_grid(feature = features, outcome = outcomes) %>%

  # apply to each combination of feature and outcome, storing the result in a list column
  rowwise() %>%
  mutate(result = list(coxph_singlerow(feature, outcome, dataset))) %>%
  
  # expands the list column into individual columns, effectively joining all the new rows together.
  unnest(result) %>% 
  
  # also add significance levels
  mutate(Significance = case_when(Pvalue < 0.05 & Pvalue > 0.001 ~ "*",
                                               Pvalue < 0.001 & Pvalue > 0.0001 ~ "**",
                                               Pvalue < 0.0001 ~ "***"))

head(sumstats_univariate_train)
```

## Make forest plot of Cox regression

Create nice labels for outcome measures
```{r}
sumstats_univariate_train$Outcome <- 
  factor(sumstats_univariate_train$Outcome,
         levels = rev(c("EVENT_VL4K","EVENT_HYBRID","EVENT_CD4")),
         labels = rev(c("Viral load > 10,000 copies/mL",
                        "Viral load >10,000 copies/mL
                        \nor >1000 increase in moving median", 
                        "CD4 count <500 cells/mm3\nmoving median")))
```
Make forest plot
```{r}
g <- ggplot(sumstats_univariate_train, aes(HR, Variable, color = Outcome, shape = Outcome)) +
   geom_pointrange(aes(xmin = Lower, xmax = Upper),
                   position = position_dodge(width = 0.5)) +
   geom_vline(xintercept = 1, linetype = 2) +
   theme_bw() +
   theme(strip.placement = "outside",
         strip.text= element_text(angle = 0, size = 6),
         strip.background = element_blank(),
         axis.text = element_text(size = 8, color="black"),
         axis.title = element_text(size = 10, color="black"),
         title = element_text(size=12, color = "black"),
         legend.text = element_text(size = 10, color="black"),
         legend.title = element_text(size = 10, color="black"),
         legend.position = "bottom",
         plot.subtitle = element_text(size = 10, color= "black", face="italic"),
         legend.key.height = unit(1.5,'cm'),
         panel.spacing = unit(0, "mm")) +
   scale_colour_manual(values=rev(c("#5ec962","orange","darkred"))) + 
     scale_x_continuous(trans='log10',limits = c(0.1,10))  +
  xlab("Hazard ratio (95% confidence interval)") + ylab("")+
  ggtitle("Baseline predictors for viral or immunological loss",
          subtitle = "Univariate Cox proportional-hazards regression")

g1 <- g + 
scale_y_discrete(
  limits = rev(c(
    "AGE", "MALESEX", "baseBMI", "ORIGIN_NONWESTERN", 
    "SMOKING_NEVER", "IDU", 
    "baseHCVever", "baseHBVever", 
    "baseCD4_sqrt", "baseVL_log10")),
  labels = rev(c(
    "Age", "Male sex", "BMI", "Non-Western\ncountry of birth", 
    "Never smoked", "Intravenous\ndrug use", 
    "History of HCV\nat baseline", "History of HBV\nat baseline", 
    "Square root\nCD4 count", "Log10 viral load")))

ggsave(paste0(outputdir, "241027_Univariate_Coxph_Train.pdf"),g1,width = 21/2, height = 29.7/2, units = "cm")
```

```{r}
sumstats_univariate_train %>% filter(Pvalue <0.05) %>% arrange(Pvalue)
```
# Train the multivariate Cox model
Remove baseBMI since there are many missing.

```{r}
# Remove those in train with missing CD4
train <- train %>% filter(!is.na(baseCD4_sqrt))

# Define features (predictors) and outcomes
features <- c("MALESEX", "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER","baseVL_log10","baseHCVever","baseHBVever","AGE","baseBMI",
              "baseCD4_sqrt")
outcomes <- c("VL4K","HYBRID", "CD4")

# Frequency of outcome happening
for (outcome in outcomes){
  event <- paste0("EVENT_", outcome)
  print(summary.factor(train[,event]))
}

# Define function to run and clean a univariate cox model for a specific feature and outcome
create_multivar_coxph <- function(features, outcome, dataset){
  # Define the formula for the cox model
  time <- paste0("TIME_", outcome)
  event <- paste0("EVENT_", outcome)
  formula_coxph <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))
  
  model <- coxph(formula_coxph, dataset%>%filter(!is.na(.data[[event]])), x = TRUE)
  
  return(model)
}

create_multivar_coxph(features, outcomes[1], train)

# note that the coxph() is so fast that I do not use the function create_multivar_coxph
## the coxph() model is created as part of obtain_auc
```

Participant counts
```{r}
train %>% filter(!is.na(baseCD4_sqrt)) %>% count()
train %>% filter(!is.na(baseCD4_sqrt)) %>% filter(!is.na(EVENT_CD4)) %>% count()

test %>% filter(!is.na(baseCD4_sqrt)) %>% count()
test %>% filter(!is.na(baseCD4_sqrt)) %>% filter(!is.na(EVENT_CD4)) %>% count()

```

## Improve cox model performance

The problem is that the model was overfitting if I include all predictors, so external testing in test results in AUCs of 0.50.

Only consider the features that were significant in univariate regression:
- VL4K & HYBRID --> MALESEX, AGE, VLlog10 and CD4sqrt
- CD4 --> MALESEX, VLlog10 and CD4sqrt, IDU, baseHCVever, ORIGIN_NONWESTERN, 
- time_since_2000 causes strong overfitting probably because interaction terms and should be discarded.
- In the RSF very consistently are the most relevant MALESEX, AGE, VLlog10 and CD4sqrt. Regardless of outcome type. Including more does not improve the model, only decreases it.
- **Conclusion: best Cox model with only MALESEX, AGE, VLlog10 and CD4sqrt
```{r}
# Define features (predictors)
features <- c("MALESEX", "AGE", 
              #"IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER", "baseHCVever", "baseHBVever",
              "baseVL_log10", "baseCD4_sqrt"
              )

# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")
# Specify time points of interest to measure AUC

outcome <- outcomes[2] 

time <- paste0("TIME_", outcome)
event <- paste0("EVENT_", outcome)
formula_model <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))

print(dim(train))
print(dim(test))

# Train the cox model
coxmodel <- create_multivar_coxph(features, outcomes[3], train)

times <- seq(2, 8) * 365
dataset <- test
survformula <- as.formula(paste("Surv(", time, ",", event, ") ~ 1"))
score_result <- Score(list("Cox" = coxmodel),
        formula = survformula,
        data = dataset%>%filter(!is.na(.data[[event]])),
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)

```

## Penalized cox models

alpha = 1 for lasso, alpha = 0 for ridge

Ridge (L2): Useful when you suspect all variables may be relevant but need regularization.
Lasso (L1): Helps with feature selection by pushing less informative coefficients to zero, simplifying the model.
```{r}
# Define features (predictors)
features <- c("MALESEX", "AGE", 
              "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER", "baseHCVever", "baseHBVever",
              "baseVL_log10", "baseCD4_sqrt"
              )

# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")
# Specify time points of interest to measure AUC

outcome <- outcomes[1] 

time <- paste0("TIME_", outcome)
event <- paste0("EVENT_", outcome)
formula_model <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))

print(dim(train))
print(dim(test))

# GLMnet from riskregression is compatible with Score(), but glmnet (lowercase) library is NOT.
# the ?glmnet is very useful for documentation on model parameters.
# alpha = 1 means elasticnet mixing parameter is lasso penalty, and alpha=0 sets it to ridge penalty
train <- train %>% filter(!is.na(baseCD4_sqrt))

penalizedmodel_lasso <- GLMnet(formula_model, data = train, alpha = 1, lambda = NULL, cv = TRUE, nfolds = 10, standardize = TRUE)

na_counts <- sapply(test, function(x) sum(is.na(x)))
print(na_counts)

times <- seq(2, 8) * 365
dataset <- test
survformula <- as.formula(paste("Surv(", time, ",", event, ") ~ 1"))
score_result <- Score(list("Penalized" = penalizedmodel_lasso),
        formula = survformula,
        data = dataset%>%filter(!is.na(.data[[event]])),
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```
## Find optimal alpha for elasticnet
```{r}
# Define a sequence of alpha values to test
alpha_values <- seq(0, 1, by = 0.1)  # Adjust step size as needed

# Initialize a data frame to store results
elasticnet_cvresults <- data.frame(alpha = numeric(), lambda_min = numeric(), cvm_min = numeric(), outcome = character())

features <- c("MALESEX", "AGE", 
              "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER", "baseHCVever", "baseHBVever",
              "baseVL_log10", "baseCD4_sqrt"
              )

for (outcome in outcomes){
  
  time <- paste0("TIME_", outcome)
  event <- paste0("EVENT_", outcome)
  formula_model <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))
  
  # Loop through each alpha value
  for (alpha in alpha_values) {
    # Perform cross-validation for each alpha
    set.seed(123)  # For reproducibility
    
    penalizedmodel_balanced <- GLMnet(formula_model, data = train, alpha = alpha, lambda = NULL, cv = TRUE, nfolds = 10, standardize = TRUE)
  
    # Store the results: best lambda and corresponding CV error
    elasticnet_cvresults <- rbind(
      elasticnet_cvresults, 
      data.frame(alpha = alpha,
                 lambda_min = penalizedmodel_balanced$fit$lambda.min,
                 cvm_min = min(penalizedmodel_balanced$fit$cvm),
                 outcome = outcome))
  }
}
elasticnet_cvresults

# Find the optimal alpha based on minimum CV error
optimal_alpha_value <- elasticnet_cvresults %>% group_by(outcome) %>% filter(cvm_min == min(cvm_min)) %>% select(alpha, outcome) %>%  pull(alpha, outcome) 

print(paste0('optimal alpha for glmnet: ',optimal_alpha_value))
```

# Gradient boosted machine

Test the code with simple random data
```{r}
# Load libraries
library(gbm)
library(riskRegression)
library(survival)  # For Surv function

# Generate sample survival data
set.seed(123)
data <- data.frame(
    time = rexp(100, rate = 0.1),          # Random survival times
    status = rbinom(100, 1, 0.5),          # Random event indicator (1 = event, 0 = censoring)
    x1 = rnorm(100),                       # Predictor 1
    x2 = rnorm(100)                        # Predictor 2
)

# Fit a GBM model with survival data
gbm_model <- gbm(
    formula = Surv(time, status) ~ x1 + x2,  # Survival formula
    data = data,
    distribution = "coxph",                    # Cox model for survival
    n.trees = 100,                           # Number of trees
    interaction.depth = 3,                   # Tree depth
    shrinkage = 0.01,                        # Learning rate
    n.minobsinnode = 10                      # Minimum observations in terminal nodes
)

# Prepare new data for prediction
new_data <- data.frame(x1 = rnorm(10), x2 = rnorm(10))

# Use Score function with the GBM survival model
score_result <- Score(
    list("GBM Model" = gbm_model), 
    formula = Surv(time, status) ~ 1,   # Survival formula for Score function
    data = data,                        # Original data with survival info
    times = c(1, 5, 10),                # Time points to evaluate the survival model
    metrics = "AUC",                  # Metric for model evaluation
    summary = "risks"                     # Summary option for integrated Brier score
)

# Print results
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```

Train the gbm model
```{r}
# Define features (predictors)
features <- c("MALESEX", "AGE",
              "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER", "baseHCVever", "baseHBVever",
              "baseVL_log10", "baseCD4_sqrt"
              )

# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")
# Specify time points of interest to measure AUC

outcome <- outcomes[1]

time <- paste0("TIME_", outcome)
event <- paste0("EVENT_", outcome)
formula_model <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))

print(dim(train))
print(dim(test))
train <- train %>% filter(!is.na(baseCD4_sqrt))

# Fit a GBM model with survival data
data <- train
data$time <- data$TIME_VL4K
data$status <- data$EVENT_VL4K
gbm_model <- gbm(
    formula = Surv(time, status) ~ MALESEX + AGE + IDU + ORIGIN_NONWESTERN + 
    SMOKING_NEVER + baseHCVever + baseHBVever + 
    baseVL_log10 + baseCD4_sqrt,  # Survival formula
    data = data,
    distribution = "coxph",                    # Cox model for survival
    n.trees = 1000,                           # Number of trees
    interaction.depth = 3,                   # Tree depth
    shrinkage = 0.01,                        # Learning rate
    n.minobsinnode = 15,                      # Minimum observations in terminal nodes
    bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
)

# Test data
test_data <- test
test_data$time <- test_data$TIME_VL4K
test_data$status <- test_data$EVENT_VL4K

# Use Score function with the GBM survival model
times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(", time, ",", event, ") ~ 1"))
score_result <- Score(
    list("GBM Model" = gbm_model), 
    formula = survformula,   # Survival formula for Score function
    data = data,                        # Original data with survival info
    times = times,                # Time points to evaluate the survival model
    metrics = "AUC",                  # Metric for model evaluation
    summary = "risks"                     # Summary option for integrated Brier score
)

# Print results
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```

Try to improve the performance of CD4 prediction
```{r}
outcome <- outcomes[3]
data <- train %>% filter(!is.na(EVENT_CD4))
data$time <- data$TIME_CD4
data$status <- data$EVENT_CD4
test_data <- test %>% filter(!is.na(EVENT_CD4))
test_data$time <- test_data$TIME_CD4
test_data$status <- test_data$EVENT_CD4

gbm_model <- gbm(
    formula = Surv(time, status) ~ MALESEX + AGE + IDU + ORIGIN_NONWESTERN + 
    SMOKING_NEVER + baseHCVever + baseHBVever + 
    baseVL_log10 + baseCD4_sqrt,  # Survival formula
    data = data,
    distribution = "coxph",                    # Cox model for survival
    n.trees = 1500,                           # Number of trees
    interaction.depth = 4,                   # Tree depth
    shrinkage = 0.05,                        # Learning rate
    n.minobsinnode = 20,                      # Minimum observations in terminal nodes
    bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
)

# Use Score function with the GBM survival model
times <- seq(2, 8) * 365
survformula <- as.formula(paste("Surv(", time, ",", event, ") ~ 1"))
score_result <- Score(
    list("GBM Model" = gbm_model), 
    formula = survformula,   # Survival formula for Score function
    data = data,                        # Original data with survival info
    times = times,                # Time points to evaluate the survival model
    metrics = "AUC",                  # Metric for model evaluation
    summary = "risks"                     # Summary option for integrated Brier score
)

# Print results
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "train"

score_result <- Score(
    list("GBM Model" = gbm_model), 
    formula = survformula,   # Survival formula for Score function
    data = test_data,                        # Original data with survival info
    times = times,                # Time points to evaluate the survival model
    metrics = "AUC",                  # Metric for model evaluation
    summary = "risks"                     # Summary option for integrated Brier score
)
AUC_overtime2 <- score_result$AUC$score
AUC_overtime2$type <- "test"

AUC_overtime <- rbind(AUC_overtime, AUC_overtime2)
print(AUC_overtime)
```

```{r}
#-----------------------------------------
# Define grid of hyperparameters to search over
n.trees.values <- c(500, 1000, 1500)
interaction.depth.values <- c(2, 3, 4)
shrinkage.values <- c(0.01, 0.05, 0.1)
n.minobsinnode.values <- c(10, 15, 20)

# Initialize variables to store the best model and best AUC score
best_auc <- -Inf
best_params <- list()

# Loop over all combinations of hyperparameters
for (n.trees in n.trees.values) {
  for (interaction.depth in interaction.depth.values) {
    for (shrinkage in shrinkage.values) {
      for (n.minobsinnode in n.minobsinnode.values) {

        # Train GBM model with the current set of hyperparameters
        gbm_model <- gbm(
          formula = Surv(time, status) ~ MALESEX + AGE + IDU + ORIGIN_NONWESTERN + 
            SMOKING_NEVER + baseHCVever + baseHBVever + 
            baseVL_log10 + baseCD4_sqrt,
          data = data,
          distribution = "coxph",
          n.trees = n.trees,
          interaction.depth = interaction.depth,
          shrinkage = shrinkage,
          n.minobsinnode = n.minobsinnode,
          bag.fraction = 0.5
        )

        # Define survival formula for Score function
        survformula <- as.formula("Surv(time, status) ~ 1")
        
        # Score model using AUC metric
        score_result <- Score(
          list("GBM Model" = gbm_model), 
          formula = survformula,
          data = test_data,
          times = seq(2, 8) * 365,
          metrics = "AUC",
          summary = "risks"
        )

        # Extract the maximum AUC from the score results
        maxauc <- max(score_result$AUC$score$AUC)

        # Update best model parameters if current AUC is the highest found
        if (maxauc > best_auc) {
          best_auc <- maxauc
          best_params <- list(
            n.trees = n.trees,
            interaction.depth = interaction.depth,
            shrinkage = shrinkage,
            n.minobsinnode = n.minobsinnode
          )
        }

        # Print current iteration results (optional)
        # cat("n.trees:", n.trees, 
        #     "interaction.depth:", interaction.depth, 
        #     "shrinkage:", shrinkage, 
        #     "n.minobsinnode:", n.minobsinnode, 
        #     "Max AUC:", maxauc, "\n")
      }
    }
  }
}

# Output the best AUC and parameters
cat("Best AUC:", best_auc, "\n")
cat("Best Parameters:\n",names(best_params), "\n", unlist(best_params)) 
```



# HAL model, highly adaptable lasso

**This is a computationally intensive model, not suited for the proteomics data.**

On the HAL model:
https://cran.r-project.org/web/packages/hal9001/hal9001.pdf

max_degree is highest order of interaction terms for basic functions
num_knots is maximum number of knots/bins for basic functions, should decrease with higher max_degree orders.
smoothness_order controls the degree of smoothness, higher is more complex but risk of overfitting. Often set to 1 or 2. 1 means main effects + twoway interactions and 2 is more complex.

The package advises as good settings:

- If smoothness_orders = 1+ and max_degree = 3, num_knots = c(100, 75, 50).

```{r}
# Define features (predictors)
# features <- c("MALESEX", "AGE", 
#               "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER", "baseHCVever", "baseHBVever",
#               "baseVL_log10", "baseCD4_sqrt"
#               )
# 
# # Define outcome measures
# outcomes <- c("VL4K","HYBRID", "CD4")
# # Specify time points of interest to measure AUC
# 
# outcome <- outcomes[1] 
# 
# time <- paste0("TIME_", outcome)
# event <- paste0("EVENT_", outcome)
# formula_model <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))
# 
# print(dim(train))
# print(dim(test))
# train <- train %>% filter(!is.na(baseCD4_sqrt))
# 
# # Measure how long this takes:
# system_time <- system.time({
#   halmodel <- Hal9001(formula_model, data = train, lambda = NULL, smoothness_orders = 1, max_degree = 3, num_knots = c(100, 75, 50))
# })
# 
# print(system_time)

#halmodel <- Hal9001(formula_model, data = train, lambda = NULL, smoothness_orders = 1, max_degree = 3, num_knots = c(100, 75, 50))
# 
# times <- seq(2, 8) * 365
# dataset <- test
# survformula <- as.formula(paste("Surv(", time, ",", event, ") ~ 1"))
# score_result <- Score(list("Penalized" = penalizedmodel_lasso),
#         formula = survformula,
#         data = dataset%>%filter(!is.na(.data[[event]])),
#         metrics = "auc",
#         summary = "risks",
#         times = times)
# AUC_overtime <- score_result$AUC$score
# AUC_overtime$type <- "test"
# print(AUC_overtime)

```


# Train the random survival forest model (RSF)
Initially without prefiltering, since there are not so many variables.
Remove baseBMI since there are many missing.

Non-parametric machine learning approach.

The rfsrc() function offers many parameters for fine-tuning:
* mtry: Number of variables randomly selected at each split
* nodesize: Minimum size of terminal nodes
* splitrule: Rule used for splitting nodes


```{r}
# Remove those in train with missing CD4 at baseline (there are no missing in the test)
train <- train %>% filter(!is.na(baseCD4_sqrt))

# Define features (predictors)
features <- c("MALESEX", "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER","baseVL_log10","baseHCVever","baseHBVever","AGE",
              "baseCD4_sqrt")

# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")
```


Tweak the model

- In survival, nodesize is minimum 15, mtry is usually 1/3 (regression) or sqrt (classification), which is 4 rounded up for our ten features

```{r}
train_rsf <- function(outcome, dataset, features,
                      nodesize = 65, mtry = 4, ntree = 1000, cvfold = 10, 
                      splitrule = "logrankscore", nsplit = 10, 
                      printresults = TRUE, tunedmodel = FALSE){
  time <- paste0("TIME_", outcome)
  event <- paste0("EVENT_", outcome)
  formula_rsf <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))
  
  # only needed for CD4 outcome:
  dataset <- dataset %>% filter(!is.na(.data[[event]]))
  
  if (tunedmodel){
    tuned_model <- tune(formula_rsf, data = train)
    print(tuned_model$optimal)
  }
  
  # train the rsf model
  rsfmodel <- rfsrc(formula_rsf, data = train, ntree = ntree, cv.fold = cvfold, nodesize = nodesize, mtry = mtry, splitrule = splitrule, nsplit = nsplit)

  # relative importance
  cv_vimp_results <- vimp(rsfmodel)
  
  if (printresults) {
    print(rsfmodel)
    plot(cv_vimp_results)
  }
  
  return(rsfmodel)
}

run_example_rsf <- FALSE
if (run_example_rsf) {
  temp <- train_rsf("VL4K", train,  features, nodesize = 15, mtry = 4, splitrule = "logrankscore")
  temp <- train_rsf("HYBRID", train, features, nodesize = 15, mtry = 4, splitrule = "logrank")
  temp <- train_rsf("CD4", train, features, nodesize = 15, mtry = 4, splitrule = "logrankscore")
}

```

Try to improve the performance of CD4 prediction
```{r}
features <- c("MALESEX", "AGE",
              "IDU", "ORIGIN_NONWESTERN", "SMOKING_NEVER","baseHCVever","baseHBVever",
              "baseCD4_sqrt","baseVL_log10")
outcome <- outcomes[2] 

time <- paste0("TIME_", outcome)
event <- paste0("EVENT_", outcome)
formula_model <- as.formula(paste("Surv(", time, ",", event, ") ~", paste(features, collapse = " + ")))

print(dim(train))
print(dim(test))

# Train the rsf model
temp <- train_rsf(outcomes[2], train, features, nodesize = 15, mtry = 4, splitrule = "logrankscore",printresults = FALSE)

times <- seq(2, 8) * 365
dataset <- test
survformula <- as.formula(paste("Surv(", time, ",", event, ") ~ 1"))
score_result <- Score(list("RSF" = temp),
        formula = survformula,
        data = dataset%>%filter(!is.na(.data[[event]])),
        metrics = "auc",
        summary = "risks",
        times = times)
AUC_overtime <- score_result$AUC$score
AUC_overtime$type <- "test"
print(AUC_overtime)
```

## Display the relevance of each predictor in the RSF model

```{r}
# filter missing data (n=8)
train <- train %>% filter(!is.na(baseCD4_sqrt))

# Define features (predictors)
features <- c("MALESEX", "AGE",
              "SMOKING_NEVER","baseHCVever","baseHBVever",
              "IDU", "ORIGIN_NONWESTERN",
              "baseCD4_sqrt", "baseVL_log10")
# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")

# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")

meandecreaseaccuracy = list()

obtain_meandecreaseaccuracy <- function(outcome){
# Rename the data in train
  train <- train %>%
    mutate(time = .data[[paste0("TIME_", outcome)]],
           status = .data[[paste0("EVENT_", outcome)]]) %>%
    ## Only relevant for CD4LOSS: some do not have the outcome
    filter(!is.na(status))

  test <- test %>%
    mutate(time = .data[[paste0("TIME_", outcome)]],
           status = .data[[paste0("EVENT_", outcome)]]) %>%
    ## Only relevant for CD4LOSS: some do not have the outcome
    filter(!is.na(status))


  # train the RSF, random survival forest
  formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))
  rsfmodel <- rfsrc(formula_model, data = train, ntree = 1000, cv.fold = 10, nodesize = 65, mtry = 4, splitrule = "logrankscore", nsplit = 10)

importance_df <- tibble(
  Outcome = outcome,
  Variable = names(vimp(rsfmodel)$importance),
  MeanDecreaseAccuracy = vimp(rsfmodel)$importance
  ) 

meandecreaseaccuracy[[outcome]] <- importance_df
return(meandecreaseaccuracy)
}

# Make a list of all results
for (outcome in outcomes){
  meandecreaseaccuracy <- obtain_meandecreaseaccuracy(outcome)
}

# Perform cleaning operations on the list
for (outcome in outcomes){
  meandecreaseaccuracy[[outcome]] <- meandecreaseaccuracy[[outcome]] %>%
    mutate(
      Outcome = factor(
        Outcome, 
        levels = c("VL4K", "HYBRID", "CD4"),
        labels = c("Virological loss", 
                   ">1000 copies/mL increase", 
                   "CD4 <500 cells/mm3")),
      Variable = factor(
        Variable,
        levels = c("AGE", "MALESEX", "baseBMI", "ORIGIN_NONWESTERN",
                   "SMOKING_NEVER", "IDU",
                   "baseHCVever", "baseHBVever",
                   "baseCD4_sqrt", "baseVL_log10"),
        labels = c("Age", "Male sex", "BMI", "Non-Western country of birth",
                  "Never smoked", "Intravenous drug use",
                  "History of HCV at baseline", "History of HBV at baseline",
                  "Square root CD4 count", "Log10 viral load"))
      )
}

# Create the ggplot
var_importance_plot <- function(outcome_index){
  i <- outcome_index
  data <- meandecreaseaccuracy[[outcomes[i]]]
  fillcolor <- c("#5ec962","orange","darkred")[i]
  g <- ggplot(data, aes(x = reorder(Variable, MeanDecreaseAccuracy, .desc = TRUE), y = MeanDecreaseAccuracy)) +
  geom_bar(stat = "identity", fill = fillcolor, width = 0.5) +
  coord_flip() +
  labs(title = "Variable Importance in Random Survival Forest",
       x = "Variables",
       y = "Mean Decrease in Accuracy") +
  theme_bw() +
  theme(strip.placement = "outside",
     strip.text = element_text(angle = 0, size = 8),
     strip.background = element_blank(),
     axis.text = element_text(size = 8, color = "black"),
     axis.title.y = element_blank(),
     axis.title.x = element_text(size = 10),
     plot.title = element_text(size = 12, face = "bold")) + 
    
    annotate("text", x = 5, y = max(data$MeanDecreaseAccuracy), 
             angle = 90, size = 8 / .pt, vjust = -1,
             label = meandecreaseaccuracy[[outcomes[i]]]$Outcome[1]) 
  
  print(g)
}

for (i in c(1:length(outcomes))){
  ggsave(paste0(outputdir, "VariableImportanceRSF_", outcomes[i], ".pdf"),var_importance_plot(i),width = 21/2, height = 29.7/6, units = "cm")

}
```



# Test performance of the models with the test datasets

Here I make a function to obtain the risk scores of the model, called obtain_scores(), which is used in another function, called obtain_auc(). The function obtain_auc() also runs the coxph() model for each outcome.
```{r}
# Remove those in train with missing CD4 at baseline (there are no missing in the test)
train <- train %>% filter(!is.na(baseCD4_sqrt))

# Define features (predictors)
features <- c("MALESEX", "AGE",
              "SMOKING_NEVER","baseHCVever","baseHBVever",
              "IDU", "ORIGIN_NONWESTERN",
              "baseCD4_sqrt", "baseVL_log10")
# Define outcome measures
outcomes <- c("VL4K","HYBRID", "CD4")

# Specify time points of interest to measure AUC
times <- seq(2, 8) * 365

# Define function to get score result
## This function will be used in the function obtain_auc below
obtain_scores <- function(models_list, dataset, survformula, traintest){
  score_result <- Score(models_list,
        formula = survformula,
        data = dataset,
        metrics = "auc",
        summary = "risks",
        times = times)
  AUC_overtime <- score_result$AUC$score
  AUC_overtime$type <- traintest
  
  return(AUC_overtime)
}

# Define function to get the AUC over time
obtain_auc <- function(outcome){
  
  # Rename the data in train
  train <- train %>%
    mutate(time = .data[[paste0("TIME_", outcome)]],
           status = .data[[paste0("EVENT_", outcome)]]) %>%
    ## Only relevant for CD4LOSS: some do not have the outcome
    filter(!is.na(status))

  test <- test %>%
    mutate(time = .data[[paste0("TIME_", outcome)]],
           status = .data[[paste0("EVENT_", outcome)]]) %>%
    ## Only relevant for CD4LOSS: some do not have the outcome
    filter(!is.na(status))

  # Check that the data is the same
  print("status" %in% colnames(train))
  print("status" %in% colnames(train))
  print(sum(train$time != train[,paste0("TIME_", outcome)]))
  print(sum(train$status != train[,paste0("EVENT_", outcome)]))
  print(sum(test$time != test[,paste0("TIME_", outcome)]))
  print(sum(test$status != test[,paste0("EVENT_", outcome)]))

  # Train the cox model
  features_cox <- c("MALESEX", "AGE","baseCD4_sqrt", "baseVL_log10")
  formula_cox <- as.formula(paste("Surv(time, status) ~", paste(features_cox, collapse = " + ")))
  coxmodel <- coxph(formula_cox, train, x = TRUE)
  
  # Define the formula for other models
  formula_model <- as.formula(paste("Surv(time, status) ~", paste(features, collapse = " + ")))
  
  # Train the elastic-net (balanced) penalized cox model
  elasticnetcox_model <- GLMnet(
    formula = formula_model,
    data = train,
    alpha = optimal_alpha_value[outcome],
    lambda = NULL,
    cv = TRUE,
    nfolds = 10,
    standardize = TRUE)

  # train the RSF, random survival forest
  rsfmodel <- rfsrc(formula_model, data = train, ntree = 1000, cv.fold = 10, nodesize = 65, mtry = 4, splitrule = "logrankscore", nsplit = 10)
  
  # Fit a GBM model with survival data
  gbm_model <- gbm(
      formula = Surv(time, status) ~ MALESEX + AGE + IDU + ORIGIN_NONWESTERN +
      SMOKING_NEVER + baseHCVever + baseHBVever +
      baseVL_log10 + baseCD4_sqrt,  # Survival formula
      data = train,
      distribution = "coxph",                    # Cox model for survival
      n.trees = 500,                           # Number of trees
      interaction.depth = 4,                   # Tree depth
      shrinkage = 0.05,                        # Learning rate
      n.minobsinnode = 15,                      # Minimum observations in terminal nodes
      bag.fraction = 0.5 # fraction of training data used for fitting, default 0.5
  )

  # Calculate AUC data over time for train and test
  survformula <- as.formula(paste("Surv(time, status) ~ 1"))
  model_list <- list("Cox" = coxmodel, 
                     "RSF" = rsfmodel,
                     "ElasticNetCox" = elasticnetcox_model,
                     "GBM" = gbm_model)
  auc_train <- obtain_scores(model_list, train, survformula, "train")
  auc_test <-  obtain_scores(model_list, test, survformula, "test")

  auc_combined <- rbind(auc_train, auc_test)
  auc_combined$outcome <- outcome

  return(auc_combined)
}

# Apply the function to each outcome
auc_data <- map_dfr(outcomes, obtain_auc)

# add nice labels for model and outcome
auc_data <- auc_data %>% mutate(
  model = factor(
    model, levels = c("Cox",
                      "RSF",
                      "ElasticNetCox",
                      "GBM"), 
    labels = c("Cox proportional hazards regression",
               "Random survival forest",
               "Elastic net cox regression",
               "Gradient boosted machines")),
  outcome = factor(outcome, 
    levels = c("VL4K", "HYBRID", "CD4"),
    labels = c("Virological loss", ">1000 copies/mL increase", "CD4 <500 cells/mm3")))

# Plot AUC over time with confidence intervals
## The plot should facetgrid on model on the x-axis and outcome on the y-axis. 
## outcome determines color and fill. type determines shape and linetype and group.
auc_plot <- ggplot(auc_data, aes(x = times, y = AUC, group = type)) +
  geom_line(aes(linetype = type, color = outcome), linewidth = 1) +
  geom_point(aes(shape = type, color = outcome), size = 2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = outcome), alpha = 0.2) +
  labs(x = "Time (years)", 
       y = "Area Under the Curve (AUC)", 
       title = "Prediction model performance over time (clinical predictors only)") +
  theme_bw() +
  scale_colour_manual(values=(c("#5ec962","orange","darkred"))) + 
  scale_fill_manual(values=(c("#5ec962","orange","darkred"))) + 
  scale_y_continuous(limits = c(0.35, 1.1),
    breaks = c(0.5, 0.75, 1.0)) +
  scale_x_continuous(limits = c(0, 9*365),
    breaks = seq(0, max(auc_data$times), by = 365),
    labels = function(x) round(x / 365, 1)
  ) +
  facet_grid(outcome ~ model) +
  theme(strip.placement = "outside",
     strip.text= element_text(angle = 0, size = 8),
     strip.background = element_blank(),
     axis.text = element_text(size = 10, color="black"),
     axis.title = element_text(size = 10, color="black", face = "bold"),
     title = element_text(size=12, color = "black"),
     legend.text = element_text(size = 10, color="black"),
     legend.title = element_blank(),
     legend.position = "bottom",
     plot.subtitle = element_text(size = 10, color= "black", face="italic"),
     #legend.key.height = unit(1.5,'cm'),
     panel.grid.minor.x = element_blank(),
     panel.spacing = unit(0, "mm")) +
  
  # keep the type legend, but remove the outcome legend
    guides(fill = "none", color = "none", shape = guide_legend(override.aes = list(size=3))) # Remove outcome legend

print(auc_plot)

ggsave(paste0(outputdir,"AUC_clinicalpredictorsonly.pdf"), plot = auc_plot, width = 210, height = 297/2, units = "mm")
```
## Create ROC curves for the models with the test datasets 

```{r}
# roc_plot <- ggplot(roc_data, aes(x = FP, y = TP, color = model, linetype = model)) +
#   geom_line(size = 1) +
#   geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
#   facet_wrap(~ times) +
#   scale_color_brewer(palette = "Set1") +
#   labs(x = "False Positive Rate", 
#        y = "True Positive Rate",
#        title = "ROC Curves for Random Survival Forest and Cox Model",
#        subtitle = paste("AUC at 60 months - RSF:", 
#                         round(roc_data$AUC[roc_data$times == 60 & roc_data$model == "RSF"], 3),
#                         "Cox:", 
#                         round(roc_data$AUC[roc_data$times == 60 & roc_data$model == "Cox"], 3))) +
#   theme_bw() +
#   coord_equal() +
#   theme(legend.position = "bottom")
# 
# print(roc_plot)

```

